---
title: "STA 210 Sticker"
author: "Maria Tackett"
date: "12/30/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE,
                      message=FALSE)
```

This document contains the code to make the sticker for [STA 210: Regression Analysis](https://www2.stat.duke.edu/courses/Spring19/sta210.001/).

```{r packages}
library(hexSticker)
library(readr)
library(tidyverse)
library(tibble)
library(showtext)
library(nnet)
```

## The Data

The [Capital Bikeshare data](https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset) is used to create the main plot for the sticker.

- `season`: 1 - Winter, 2 - Spring, 3 - Summer 4 - Fall 
- `atemp`: feeling temperature &#247; 50 (in degrees Celsius)
<br> 

```{r data}
bikeshare <- read_csv("https://raw.githubusercontent.com/matackett/data/master/capital-bikeshare.csv")    
bikeshare <- bikeshare %>%
  mutate(season = case_when(
    season==1 ~ "Winter",
    season==2 ~ "Spring",
    season==3 ~ "Summer",
    season==4 ~ "Fall"
  )) %>% 
  select(season,atemp)
```

## Fit Model

A multinomial logistic regression model is used to create the main plot for the sticker. The response variable is `season` and the predictor variable is `atemp`. 

```{r multi-logistic}
# fit multinomial logistic model 
# add predicted probabilities to bikeshare data
m <- multinom(season ~  atemp,data=bikeshare)
pred <- as.data.frame(predict.glm(m,type="response"))
bikeshare <- bind_cols(bikeshare,pred)
```

## Make Sticker

```{r plot}
p <- ggplot(data=bikeshare,aes(x=atemp)) + 
  geom_line(aes(y=Winter),color="#00BFC4") +
  geom_line(aes(y=Spring),color="#F8766D") +
  geom_line(aes(y=Summer),color="#C77CFF") +
  geom_line(aes(y=Fall),color="#7CAE00") +
  theme_minimal() +
  theme(axis.title=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank())
```

```{r add-font}
font_add_google("Open Sans", "open")
```

```{r make-sticker}
sticker(p, package="STA 210",p_color="#00797C", p_family="open", p_size=7.5, s_x=1, s_y=0.75, s_width=1.2, s_height=1, h_fill = "#FFFFFF", 
        h_color="#00797C", h_size =0.8,
        filename="sta210_sticker.png")
```