---
title: "Easy Website Background Using ggplot2"
author: "Dr. Maria Tackett"
date: "12.17.2018"
output: 
  html_document: 
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE,
                      message=FALSE)
```

## Introduction 

This document contains the code used to make the background image for the [STA 199 Spring 2019](https://www2.stat.duke.edu/courses/Spring19/sta199.001/) course website. This background can be used to start a discussion with introductory students about data visualization and the information conveyed using various plot types.



## The Data

The [Capital Bikeshare data](https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset) is used to create this background. Each plot uses a combination of the following variables: 

- `cnt`: number of bike rentals
- `season`: 1 - Winter, 2 - Spring, 3 - Summer 4 - Fall 
- `atemp`: feeling temperature &#247; 50 (in degrees Celsius)
<br> 

```{r load-packages}
library(tidyverse)
library(readr)
library(cowplot)
```

```{r data}
bikeshare <- read_csv("https://raw.githubusercontent.com/matackett/data/master/capital-bikeshare.csv")                          
```


```{r change-season}
bikeshare <- bikeshare %>%
  mutate(season = case_when(
    season==1 ~ "Winter",
    season==2 ~ "Spring",
    season==3 ~ "Summer",
    season==4 ~ "Fall"
  )) %>% 
  select(season,atemp,cnt)
```

## The Plots

A few notes about the plots: 

- The plots used in this background are scatterplot, density plot, countour plot, linear smoothing, side-by-side boxplots, and area plots. See the [ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf) for details and more plot options.

-  The titles, axes, and legend have been removed for each plot. These elements can be easily added by removing the `theme()` modifier. 

```{r make-plots}
alpha.level = 0.5 

p1 <- ggplot(data=bikeshare,aes(x=atemp,y=cnt,color=season)) +
  geom_point(alpha=alpha.level) +
  theme_void() +
  theme(legend.position="none",
        axis.title=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank()
      )

p2 <- ggplot(data=bikeshare,aes(x=cnt,fill=season)) +
  geom_density(alpha=alpha.level) +
  theme_void() +
  theme(legend.position="none",
        axis.title=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank()
      )

p3 <- ggplot(data=bikeshare,aes(x=cnt,y=atemp,color=season)) +
  geom_density2d() +
  theme_void() +
  theme(legend.position="none",
        axis.title=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank()
      )

p4 <- ggplot(data=bikeshare,aes(x=atemp,y=cnt,color=season)) +
  geom_smooth(se=FALSE) +
  theme_void() +
  theme(legend.position="none",
        axis.title=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank()
      )

p5 <- ggplot(data=bikeshare,aes(x=season,y=cnt,fill=season)) +
  geom_boxplot(alpha=alpha.level) +
  theme_void() +
  theme(legend.position="none",
        axis.title=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank()
      )

p6 <- ggplot(data=bikeshare,aes(x=cnt,fill=season)) +
  geom_area(stat="bin",alpha=alpha.level,color="black") +
  theme_void() +
  theme(legend.position="none",
        axis.title=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank()
      )
plot_grid(p1,p2,p3,p4,p5,p6,ncol=3)
```

```{r save-plot}
# modify file location
ggsave("static/img/bikeshare-plots.png",scale=2)
```


