---
title: "Why Use Regression?"
author: "Dr. Maria Tackett"
date: "01.14.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")

options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3.75,
	fig.width = 6.25,
	message = FALSE,
	warning = FALSE
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Questions about Course

---

- **I do not have a lot of prior experience using R.**
    - That's OK! 
    - RStudio Cloud primers to help you get up to speed. 
    - [Rfun - Intro to R](https://duke.libcal.com/event/4799048) workshop by Data and Visualization Services
    
- **I didn't know a lot of answers on the pretest.**
    - That's OK!
    - We will go over the concepts covered in the pretest. 
    - The only prereq for this course is an introductory statistics class at Duke.
    
---

## What is Regression Analysis?

"In statistical modeling, regression analysis is a set of statistical processes for estimating the relationships among variables. It includes many techniques for modeling and analyzing several variables, when <font class="vocab">**the focus is on the relationship between a dependent variable and one or more independent variables (or 'predictors')**</font>. More specifically, regression analysis helps one understand how the typical value of the dependent variable (or 'criterion variable') changes when any one of the independent variables is varied, while the other independent variables are held fixed."

.pull-right[
[- Wikipedia](https://en.wikipedia.org/wiki/Regression_analysis)
]

---

## What's this course about? 

- Parametric & Supervised Learning 

---

## What's not in this course?

- Unsupervised 
- Nonparametric methods

You'll get to these in STA 325: Machine Learning & Data Mining

---





---

## Before next class

- Complete [Reading 01](https://www2.stat.duke.edu/courses/Spring19/sta210.001/reading/reading-01.html)

- Complete "getting to know you survey" and pretest if you haven't already


  
