---
title: "Multiple Comparisons"
author: "Dr. Maria Tackett"
date: "09.18.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE
)
```

class: middle, center

### [Click for PDF of slides](06-mult-comparisons.pdf)

---

### Announcements

- HW 01 **due TODAY at 11:59p**

- Reading 03 for Monday

- HW 02 due Wednesday, 9/25 at 11:59p

---

### Today's Agenda 

- Multiple comparisions 

- Introducing multiple linear regression 

---

## R Packages used in the notes

```{r}
library(tidyverse)
library(knitr)
library(broom)
```

---

class: middle, center

## Multiple Comparisons

---

### After ANOVA: Individual Group Means

- Suppose you conduct an ANOVA and conclude that at least one group mean has a different mean response value. 

- The next question you want to answer is **which group?**

--

- One way to answer this question is to compare the estimated means for each group, accounting for the random variability we'd naturally expect

--

- Since we've assumed the variance is the same for all groups, we can use a pooled standard error with $n-K$ degrees of freedom to calculate the confidence

.alert[
$$\bar{y}_i \pm t^* \times \frac{s_P}{\sqrt{n_i}}$$
where $s_P$ is the pooled standard deviation
]

---

 
### After ANOVA: Difference in Means

- We can also estimate the difference in two means, $\mu_1-\mu_2$ for each pair of groups

.alert[
$$(\bar{y}_1-\bar{y}_2) \pm t^* \times s_P\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}$$
where $s_P$ is the pooled standard deviation
]

- If we have $K$ groups, we will make ${K \choose 2} = K(K-1)/2$ such comparisons
    - Ex: If we have 6 groups, we'll make ${6 \choose 2} = 6(6-1)/2 = 15$ comparisons

---

### Multiple Comparisons

- When making multiple comparisons, there is a higher chance that a Type I error will occur, e.g. conclude that there is a significant difference between two groups even when there is not


- **At a Minimum**: When calculating multiple confidence intervals or conducting multiple hypothesis tests to compare means, you should clearly state how many CIs and/or tests you computed.


- **Good practice**: Account for the number of comparisons being made in the analysis 
    - We will discuss one method: <font class = "vocab">Bonferroni correction</font>

---

### Confidence levels

- <font class="vocab">Individual confidence level: </font> success rate of a procedure for calculating a <u>single</u> confidence interval


--
- <font class="vocab">Familywise confidence level: </font> success rate of a procedure for calculating a <u>family</u> of confidence intervals 
  + "success": all intervals in the family capture their parameters


--
- <font class="vocab">Issue: </font>There is an increased chance of making at least one error when calculating multiple confidence intervals
  + The same is true when conducting multiple hypothesis tests


---

### Bonferroni correction

- <font class="vocab">Goal: </font> Achieve at least $100(1-\alpha)$% familywise confidence level for $\mathcal{C}$ confidence intervals 
    - Where $\alpha$ is the significance level for the corresponding two-sided hypothesis test

--

- Calculate each of the $k$ confidence intervals at a $100(1-\frac{\alpha}{\mathcal{C}})$% confidence level
    - When there are $K$ groups, there are $\mathcal{C}=\frac{K(K-1)}{2}$ pairs of means being compared
--

- **Notes**: 
    + The exact familywise confidence level is not easily predictable. This partially depends on the level of dependence between the intervals. 
    + Bonferroni correction is sometimes too conservative, i.e don't reject $H_0$ as much as you should

---

### Population Density in the Midwest

- There are 5 groups (states) in the `midwest` data, so we will do ${5 \choose 2} = 10$ comparisons. 

- If we want a familywise confidence level of 95%, then we should use a $(1 - 0.05/10)\times 100 = 99.5$% confidence level for each pairwise comparison

---

### Pairwise CI

```{r echo = F}
midwest <- midwest %>% mutate(log_popdensity = log(popdensity))
```

.small[
```{r}
library(pairwiseCI)
pairwiseCI(log_popdensity ~ state, data = midwest, 
           method = "Param.diff", conf.level = 0.995, var.equal = TRUE)
```
]

---

### Pairwise CI 

```{r echo = F}
library(pairwiseCI)
pairwiseCI(log_popdensity ~ state, data = midwest, 
           method = "Param.diff", conf.level = 0.995, var.equal = TRUE) %>%
  kable(format = "markdown", digits = 3)
```

