---
title: "Multiple Linear Regression"
subtitle: "Special Predictors"
author: "Prof. Maria Tackett"
date: "02.10.20"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3.5,
	message = FALSE,
	warning = FALSE, 
	dpi = 300
)
```

class: middle, center

### [Click for PDF of slides](07-mlr-special-predictors.pdf)

---

### Announcements

- Lab 04 **due tomorrow at 11:59p**
    - pdf of instructions in GitHub repo

- HW 02 **due Wed, Feb 12 at 11:59p**
  - pdf of instructions in GitHub repo
  
- [Reading for today & Wednesday](https://www2.stat.duke.edu/courses/Spring20/sta210.001/reading/reading-05.html)


---

### Today's agenda 

- Inference for regression coefficients

- Prediction

- Quick math details

- Special predictors

---
 
### R packages

```{r echo = F}
library(tidyverse)
library(knitr)
library(broom)
library(cowplot) # use plot_grid function
```

---

### House prices in Levittown (sec. 1.4)
 
- Public data on the sales of 85 homes in Levittown, NY from June 2010 to May 2011

- Levittown was built right after WWI and was the first planned suburban community built using mass production techniques 

<br>

.alert[
**Questions**: 
- What is the relationship between the characteristics of a house in Levittown and its sale price? 
- Given its characteristics, what is the expected sale price of a house in Levittown?  
]

---

### Data

```{r,echo=F}
homes <- read_csv("data/homeprices.csv")
```

```{r}
glimpse(homes)
```

---

### Variables

**Predictors**
- .vocab[`bedrooms`]: Number of bedrooms
- .vocab[`bathrooms`]: Number of bathrooms
- .vocab[`living_area`]: Total living area of the house (in square feet)
- .vocab[`lot_size`]: Total area of the lot (in square feet)
- .vocab[`year_built`]: Year the house was built
- .vocab[`property_tax`]: Annual property taxes (in U.S. dollars)

**Response**
- .vocab[`sale_price`]: Sales price (in U.S. dollars)

---

### EDA: Response variable

```{r echo = F}
ggplot(data = homes, aes(x = sale_price)) +
  geom_histogram() + 
  labs(title = "Distribution of Sale Price", 
       x = "Sale Price (in US Dollars)")
```


---

### EDA: Predictor variables 

```{r echo = F}
p1 <- ggplot(data = homes, aes(x = bedrooms)) +
  geom_histogram()

p2 <- ggplot(data = homes, aes(x = bathrooms)) +
  geom_histogram()

p3 <- ggplot(data = homes, aes(x = living_area)) +
  geom_histogram()

p4 <- ggplot(data = homes, aes(x = lot_size)) +
  geom_histogram()

p5 <- ggplot(data = homes, aes(x = year_built)) +
  geom_histogram()

p6 <- ggplot(data = homes, aes(x = property_tax)) +
  geom_histogram()

plot_grid(p1, p2, p3, p4, p5, p6, ncol = 3)
```

---

### EDA: Response vs. Predictors

```{r echo = F}
p1 <- ggplot(data = homes, aes(x = bedrooms, y = sale_price)) +
  geom_point()

p2 <- ggplot(data = homes, aes(x = bathrooms, y = sale_price)) +
  geom_point()

p3 <- ggplot(data = homes, aes(x = living_area, y = sale_price)) +
  geom_point()

p4 <- ggplot(data = homes, aes(x = lot_size, y = sale_price)) +
  geom_point()

p5 <- ggplot(data = homes, aes(x = year_built, y = sale_price)) +
  geom_point()

p6 <- ggplot(data = homes, aes(x = property_tax, y = sale_price)) +
  geom_point()

plot_grid(p1, p2, p3, p4, p5, p6, ncol = 3)
```

---

### Regression Output

.small[
```{r echo = F}
price_model <- lm(sale_price ~ bedrooms + bathrooms 
                  + living_area + lot_size 
                  + year_built + property_tax, 
                  data = homes)
```

```{r echo = F}
tidy(price_model, conf.int = TRUE) %>%
  kable(format = "markdown", digits = 3)
```
]

---

### Interpreting $\hat{\beta}_j$

- An estimated coefficient $\hat{\beta}_j$ is the expected change in $y$ to change when $x_j$ increases by one unit **<u>holding the values of all other predictor variables constant</u>**.

--

<br>

- *Example:* The estimated coefficient for **`living_area`** is 65.90. This means for each additional square foot of living area, we expect the sale price of a house in Levittown, NY to increase by $65.90, on average, holding all other predictor variables constant.

---

### Hypothesis Tests for $\hat{\beta}_j$

- We want to test whether a particular coefficient has a value of 0 in the population, given all other variables in the model: 

.alert[
$$\begin{aligned}&H_0: \beta_j = 0 \\ &H_a: \beta_j \neq 0\end{aligned}$$
]

- The test statistic reported in R is the following: 

$$\text{test statistic} = t =  \frac{\hat{\beta}_j - 0}{SE(\hat{\beta}_j)}$$

- Calculate the p-value using the $t$ distribution with $n - p - 1$ degrees of freedom, where $p$ is the number of terms in the model (not including the intercept).

---

### Confidence Interval for $\beta_j$

.alert[
The $C%$ confidence interval for $\beta_j$ 
$$\hat{\beta}_j \pm t^* SE(\hat{\beta}_j)$$
where $t^*$ follows a $t$ distribution with with $(n - p - 1)$ degrees of freedom
]

- **General Interpretation**: We are $C%$ confident that the interval LB to UB contains the population coefficient of $x_j$. Therefore, for every one unit increase in $x_j$, we expect $y$ to change by LB to UB units, holding all else constant. 

---

### Confidence interval for `living_area`

```{r,echo=F, eval = F}
tidy(price_model, conf.int = TRUE) %>%
  kable(format = "markdown", digits = 3)
```

.question[
Interpret the 95% confidence interval for the coefficient of `living_area`.
]


---

### Caution: Large sample sizes


If the sample size is large enough, the test will likely result in rejecting $H_0: \beta_j=0$ even $x_j$ has a very small effect on $y$

- Consider the <font class="vocab">practical significance</font> of the result not just the statistical significance 

- Use the confidence interval to draw conclusions instead of p-values
 
---

### Caution: Small sample sizes

If the sample size is small, there may not be enough evidence to reject $H_0: \beta_j=0$

- When you fail to reject the null hypothesis, **DON'T** immediately conclude that the variable has no association with the response. 
  
- There may be a linear association that is just not strong enough to detect given your data, or there may be a non-linear association.
    

---

### Prediction

- We calculate predictions the same as with simple linear regression 

- **Example:** What is the predicted sale price for a house in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with $6306 in property taxes?


```{r}
-7148818.957 - 12291.011 * 3 + 51699.236 * 1 + 
  65.903 * 1050 - 0.897 * 6000 + 3760.898 * 1948 + 1.476 * 6306
```

--

The predicted sale price for a house in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with $6306 in property taxes is **$265,360**.

---

### Intervals for predictions

- Just like with simple linear regression, we can use the .vocab[`predict`] function in R to calculate the appropriate intervals for our predicted values 

```{r eval = F}
x0 <- data.frame(bedrooms = 3, bathrooms = 1, living_area = 1050, 
                 lot_size = 6000, year_built = 1948, 
                 property_tax = 6306)
predict(price_model, x0, interval = "prediction")
```

.question[

- Go to http://bit.ly/sta210-sp20-pred and use the model to answer the questions

- Use **NetId@duke.edu** for your email address.

- You are welcome (and encouraged!) to discuss these questions with 1 - 2 people around you, but **each person** must submit a response.
]

```{r echo = F}
library(countdown)
countdown(minutes = 3, seconds = 0, update_every = 1, warn_when = 30)
```

---

### Cautions

- .vocab3[Do not extrapolate!] Because there are multiple explanatory variables, you can extrapolation in many ways

--

- The multiple regression model only shows .vocab3[association, not causality]
  + To show causality, you must have a carefully designed experiment or carefully account for confounding variables in an observational study 
 
---

class: middle, center

## Math details

---

### Regression Model 

- The multiple linear regression model assumes 

.alert[
$$y|x_1, x_2,  \ldots, x_p \sim N(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p, \sigma^2)$$
]

--

- For a given observation $(x_{i1}, x_{i2}, \ldots, x_{ip}, y_i)$, we can rewrite the previous statement as 

.alert[
$$y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} + \epsilon_{i} \hspace{10mm} \epsilon_i \sim N(0,\sigma^2)$$
]
---

### Estimating $\sigma^2$

- For a given observation $(x_{i1}, x_{i2}, \ldots,x_{ip}, y_i)$ the residual is 

.alert[
$$e_i = y_{i} - (\hat{\beta}_0 + \hat{\beta}_1 x_{i1} + \hat{\beta}_{2} x_{i2} + \dots + \hat{\beta}_p x_{ip})$$
]

--

- The estimated value of the regression variance , $\sigma^2$, is 

.alert[
$$\hat{\sigma}^2  = \frac{RSS}{n-p-1} = \frac{\sum_{i=1}^ne_i^2}{n-p-1}$$
]

---

### Estimating Coefficients 

- One way to estimate the coefficients is by taking partial derivatives of the formula

$$\sum_{i=1}^n e_i^2 = \sum_{i=1}^{n}[y_i - (\beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} \dots + \beta_p x_{ip})]^2$$

--

- This produces messy formulas, so instead we can use matrix notation for multiple linear regression and estimate the coefficients using rules from linear algebra. 
    - For more details, see Section 1.2 of the textbook and the supplemental notes [Matrix Notation for Multiple Linear Regression](https://github.com/sta210-sp20/supplemental-notes/blob/master/regression-basics-matrix.pdf)
    - **Note:** You are **<u>not</u>** required to know matrix notation for MLR in this class

---

class: middle, center

## Special Predictors

---

### Interpreting the Intercept 

```{r echo=FALSE}
kable(tidy(price_model),format="html",digits=3)
```

.question[
- Interpret the intercept. 

- Is this interpretation meaningful? Why or why not?
]

---


### Mean-Centered Variables

- To have a meaningful interpretation of the intercept, use **mean-centered** predictor variables in the model (quantitative predictors only)

- A <font class = "vocab">mean-centered variable</font> is calculated by subtracting the mean from each value of the variable, i.e. $$x_{ip} - \bar{x}_{.p}$$

- Now the intercept is interpreted as the expected value of the response at the mean value of all quantitative predictors

---

### Salary: Mean-Centered Variables

```{r}
homes <- homes %>%
  mutate(bedroomsCent = bedrooms - mean(bedrooms), 
         bathroomsCent = bathrooms - mean(bathrooms), 
         living_areaCent = living_area - mean(living_area), 
         lot_sizeCent = lot_size - mean(lot_size), 
         year_builtCent = year_built - mean(year_built), 
         property_taxCent = property_tax - mean(property_tax))
```

.pull-left[
```{r,echo=F}
ggplot(data=homes, aes(x = living_area)) + 
  geom_histogram(fill ="steelblue2",color = "black") + 
labs(title = "Living Area", x = "Living Area (in sq ft)")
```
]
.pull-right[
```{r,echo=F}
ggplot(data=homes, aes(x = living_areaCent)) + 
  geom_histogram(fill ="green",color = "black") + 
labs(title = "Mean-Centered Living Area", x = "Mean-Centered Living Area (in sq ft)")
```
]


---

### In-class exercise

Below is the original model: 

.small[
```{r echo = F}
tidy(price_model) %>%
  kable(format = "markdown", digits = 3)
```
]

.pull-left[
.question[

- Go to http://bit.ly/sta210-sp20-mean-center and describe how the model would change if mean-centere variables were used instead.
]
]

```{r echo = F}
library(countdown)
countdown(minutes = 3, seconds = 0, update_every = 1, warn_when = 30)
```
---

### How model changes with mean-centered variables

---

### Indicator (dummy) variables


- Suppose there is a categorical variable with $k$ levels (categories)

- Make $k$ indicator variables (also known as dummy variables)

- Use $k-1$ of the indicator variables in the model
    - Can't uniquely estimate all $k$ variables at once if the intercept is in the model
    
- Level that doesn't have a variable in the model is called the <font class="vocab3">baseline</font>

- Coefficients interpreted as the change in the mean of the response over the baseline

---

### Indicator variables when $k > 2$

Suppose we create a new variable called **`decade_built`** that is the decade the house is built 

.small[
```{r}
homes <- homes %>% 
  mutate(decade_built = case_when(
    year_built %in% c(1948, 1949) ~ "Forties", 
    year_built %in% c(1961, 1962) ~ "Sixties",
    TRUE ~ "Fifties"
  ), 
  decade_built = factor(decade_built, 
                        levels = c("Forties", "Fifties", "Sixties"))
  )
```
]

---

### `year_built` and `decade_built`

```{r echo = F}
p1 <- ggplot(data = homes, aes(x = year_built)) + 
  geom_histogram() + 
  labs(title = "Year Built")

p2 <- ggplot(data = homes, aes(x = decade_built)) + 
  geom_bar() + 
  labs(title = "Decade Built")

plot_grid(p1, p2, ncol = 2)
```

---

### Model with categorical predictor

- Let's fit the model with `decade_built` instead of `year_built`

```{r echo = F}
price_model <- lm(sale_price ~ bedrooms + bathrooms + living_area + lot_size + property_tax + decade_built, 
                  data = homes)
price_model %>% 
  tidy() %>% 
  kable(format = "markdown", digits = 3)
```



---

### Interaction Terms

- **Case**: Relationship of the predictor variable with the response depends on the value of another predictor variable
  + This is an .vocab[interaction effect]
  
- Create a new interaction variable that is one predictor variable times the other in the interaction 

-  **Good Practice**: When including an interaction term, also *include the associated <u>main effects</u>* (each predictor variable on its own) even if their coefficients are not statistically significant

---

### Interaction: decase_built * bathrooms

.small[
```{r}
price_model <- lm(sale_price ~ bedrooms + bathrooms + living_area +
                    lot_size + property_tax + decade_built + 
                    decade_built * bathrooms, #<<
                  data = homes)

```
]

```{r echo =F}
price_model %>% 
  tidy() %>% 
  kable(format = "markdown", digits = 3)
```



