---
title: "Multiple Linear Regression"
subtitle: "Special Predictors"
author: "Prof. Maria Tackett"
date: "02.10.20"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3.5,
	message = FALSE,
	warning = FALSE, 
	dpi = 300
)
```

class: middle, center

### [Click for PDF of slides](07-mlr-special-predictors.pdf)

---

### Announcements

- Lab 04 **due tomorrow at 11:59p**
    - pdf of instructions in GitHub repo

- HW 02 **due Wed, Feb 12 at 11:59p**
  - pdf of instructions in GitHub repo
  
- [Reading for today & Wednesday](https://www2.stat.duke.edu/courses/Spring20/sta210.001/reading/reading-05.html)

- StatSci Majors Union: Careers at Research in Sports Analytics
  - Tuesday at 7p
  - Old Chem lobby

---

### Today's agenda 

- Inference for regression coefficients

- Prediction

- Quick math details

- Special predictors

```{r echo = F}
library(tidyverse)
library(knitr)
library(broom)
library(cowplot) # use plot_grid function
```

---

### House prices in Levittown (sec. 1.4)
 
- Public data on the sales of 85 homes in Levittown, NY from June 2010 to May 2011

- Levittown was built right after WWI and was the first planned suburban community built using mass production techniques 

<br>

.alert[
**Questions**: 
- What is the relationship between the characteristics of a house in Levittown and its sale price? 
- Given its characteristics, what is the expected sale price of a house in Levittown?  
]

---

### Data

```{r,echo=F}
homes <- read_csv("data/homeprices.csv")
```

```{r}
glimpse(homes)
```

---

### Variables

**Predictors**
- .vocab[`bedrooms`]: Number of bedrooms
- .vocab[`bathrooms`]: Number of bathrooms
- .vocab[`living_area`]: Total living area of the house (in square feet)
- .vocab[`lot_size`]: Total area of the lot (in square feet)
- .vocab[`year_built`]: Year the house was built
- .vocab[`property_tax`]: Annual property taxes (in U.S. dollars)

**Response**
- .vocab[`sale_price`]: Sales price (in U.S. dollars)

---

### EDA: Response variable

```{r echo = F}
ggplot(data = homes, aes(x = sale_price)) +
  geom_histogram() + 
  labs(title = "Distribution of Sale Price", 
       x = "Sale Price (in US Dollars)")
```


---

### EDA: Predictor variables 

```{r echo = F}
p1 <- ggplot(data = homes, aes(x = bedrooms)) +
  geom_histogram()

p2 <- ggplot(data = homes, aes(x = bathrooms)) +
  geom_histogram()

p3 <- ggplot(data = homes, aes(x = living_area)) +
  geom_histogram()

p4 <- ggplot(data = homes, aes(x = lot_size)) +
  geom_histogram()

p5 <- ggplot(data = homes, aes(x = year_built)) +
  geom_histogram()

p6 <- ggplot(data = homes, aes(x = property_tax)) +
  geom_histogram()

plot_grid(p1, p2, p3, p4, p5, p6, ncol = 3)
```

---

### EDA: Response vs. Predictors

```{r echo = F}
p1 <- ggplot(data = homes, aes(x = bedrooms, y = sale_price)) +
  geom_point()

p2 <- ggplot(data = homes, aes(x = bathrooms, y = sale_price)) +
  geom_point()

p3 <- ggplot(data = homes, aes(x = living_area, y = sale_price)) +
  geom_point()

p4 <- ggplot(data = homes, aes(x = lot_size, y = sale_price)) +
  geom_point()

p5 <- ggplot(data = homes, aes(x = year_built, y = sale_price)) +
  geom_point()

p6 <- ggplot(data = homes, aes(x = property_tax, y = sale_price)) +
  geom_point()

plot_grid(p1, p2, p3, p4, p5, p6, ncol = 3)
```

---

### Regression Output

.small[
```{r echo = F}
price_model <- lm(sale_price ~ bedrooms + bathrooms 
                  + living_area + lot_size 
                  + year_built + property_tax, 
                  data = homes)
```

```{r echo = F}
tidy(price_model, conf.int = TRUE) %>%
  kable(format = "markdown", digits = 3)
```
]

---

### Interpreting $\hat{\beta}_j$

- An estimated coefficient $\hat{\beta}_j$ is the expected change in $y$ to change when $x_j$ increases by one unit **<u>holding the values of all other predictor variables constant</u>**.

--

<br>

- *Example:* The estimated coefficient for **`living_area`** is 65.90. This means for each additional square foot of living area, we expect the sale price of a house in Levittown, NY to increase by $65.90, on average, holding all other predictor variables constant.

---

### Hypothesis Tests for $\hat{\beta}_j$

- We want to test whether a particular coefficient has a value of 0 in the population, given all other variables in the model: 

.alert[
$$\begin{aligned}&H_0: \beta_j = 0 \\ &H_a: \beta_j \neq 0\end{aligned}$$
]

- The test statistic reported in R is the following: 

$$\text{test statistic} = t =  \frac{\hat{\beta}_j - 0}{SE(\hat{\beta}_j)}$$

- Calculate the p-value using the $t$ distribution with $n - p - 1$ degrees of freedom, where $p$ is the number of terms in the model (not including the intercept).

---

### Confidence Interval for $\beta_j$

.alert[
The $C%$ confidence interval for $\beta_j$ 
$$\hat{\beta}_j \pm t^* SE(\hat{\beta}_j)$$
where $t^*$ follows a $t$ distribution with with $(n - p - 1)$ degrees of freedom
]

- **General Interpretation**: We are $C%$ confident that the interval LB to UB contains the population coefficient of $x_j$. Therefore, for every one unit increase in $x_j$, we expect $y$ to change by LB to UB units, holding all else constant. 

---

### Confidence interval for `living_area`

```{r,echo=F, eval = F}
tidy(price_model, conf.int = TRUE) %>%
  kable(format = "markdown", digits = 3)
```

.question[
Interpret the 95% confidence interval for the coefficient of `living_area`.
]


---

### Caution: Large sample sizes


If the sample size is large enough, the test will likely result in rejecting $H_0: \beta_j=0$ even $x_j$ has a very small effect on $y$

- Consider the <font class="vocab">practical significance</font> of the result not just the statistical significance 

- Use the confidence interval to draw conclusions instead of p-values
 
---

### Caution: Small sample sizes

If the sample size is small, there may not be enough evidence to reject $H_0: \beta_j=0$

- When you fail to reject the null hypothesis, **DON'T** immediately conclude that the variable has no association with the response. 
  
- There may be a linear association that is just not strong enough to detect given your data, or there may be a non-linear association.
    

---

### Prediction

- We calculate predictions the same as with simple linear regression 

- **Example:** What is the predicted sale price for a house in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with $6306 in property taxes?


```{r}
-7148818.957 - 12291.011 * 3 + 51699.236 * 1 + 
  65.903 * 1050 - 0.897 * 6000 + 3760.898 * 1948 + 1.476 * 6306
```

--

The predicted sale price for a house in Levittown, NY with 3 bedrooms, 1 bathroom, 1050 square feet of living area, 6000 square foot lot size, built in 1948 with $6306 in property taxes is **$265,360**.

---

### Intervals for predictions

- Predictions have uncertainity just like any other quantity that is estimated, so we so we want to report the appropriate interval along with the predicted value. 

.question[

- Go to http://bit.ly/sta210-sp20-pred and use the model to answer the questions

- Use **NetId@duke.edu** for your email address.

- You are welcome (and encouraged!) to discuss these questions with 1 - 2 people around you, but **each person** must submit a response.
]

```{r echo = F}
library(countdown)
countdown(minutes = 3, seconds = 0, update_every = 1, warn_when = 30)
```

---

### Intervals for predcitions 

```{r}
x0 <- data.frame(bedrooms = 3, bathrooms = 1, living_area = 1050, 
                 lot_size = 6000, year_built = 1948, 
                 property_tax = 6306)
```
--

Predict the **mean** response for the **subset** of observations that have the given characteristics: 

```{r}
predict(price_model, x0, interval = "confidence")
```

--

Predict the response for an **individual** observation with the given characteristics: 

```{r}
predict(price_model, x0, interval = "prediction")
```

---

### Cautions

- .vocab3[Do not extrapolate!] Because there are multiple explanatory variables, you can extrapolation in many ways

--

- The multiple regression model only shows .vocab3[association, not causality]
  + To show causality, you must have a carefully designed experiment or carefully account for confounding variables in an observational study 
 
---

class: middle, center

## Math details

---

### Regression Model 

- The multiple linear regression model assumes 

.alert[
$$y|x_1, x_2,  \ldots, x_p \sim N(\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \dots + \beta_p x_p, \sigma^2)$$
]

--

- For a given observation $(x_{i1}, x_{i2}, \ldots, x_{ip}, y_i)$, we can rewrite the previous statement as 

.alert[
$$y_i = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2} + \dots + \beta_p x_{ip} + \epsilon_{i} \hspace{10mm} \epsilon_i \sim N(0,\sigma^2)$$
]
---

### Estimating $\sigma^2$

- For a given observation $(x_{i1}, x_{i2}, \ldots,x_{ip}, y_i)$ the residual is 

.alert[
$$e_i = y_{i} - (\hat{\beta}_0 + \hat{\beta}_1 x_{i1} + \hat{\beta}_{2} x_{i2} + \dots + \hat{\beta}_p x_{ip})$$
]

--

- The estimated value of the regression variance , $\sigma^2$, is 

.alert[
$$\hat{\sigma}^2  = \frac{RSS}{n-p-1} = \frac{\sum_{i=1}^ne_i^2}{n-p-1}$$
]

---

### Estimating Coefficients 

- One way to estimate the coefficients is by taking partial derivatives of the formula

$$\sum_{i=1}^n e_i^2 = \sum_{i=1}^{n}[y_{i} - (\hat{\beta}_0 + \hat{\beta}_1 x_{i1} + \hat{\beta}_{2} x_{i2} + \dots + \hat{\beta}_p x_{ip})]^2$$

--

- This produces messy formulas, so instead we can use matrix notation for multiple linear regression and estimate the coefficients using rules from linear algebra. 
    - For more details, see Section 1.2 of the textbook and the supplemental notes [Matrix Notation for Multiple Linear Regression](https://github.com/sta210-sp20/supplemental-notes/blob/master/regression-basics-matrix.pdf)
    - **Note:** You are **<u>not</u>** required to know matrix notation for MLR in this class

