---
title: "Multiple Linear Regression"
subtitle: "Interactions & Transformations"
author: "Prof. Maria Tackett"
date: "02.17.20"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 4,
	message = FALSE,
	warning = FALSE, 
	dpi = 300
)
```


class: middle, center

### [Click for PDF of slides](09-transformations.pdf)

---

### Announcements

- Team Feedback #1 **due Wed, Feb 19 at 11:59p**
    - Check for email from Teammates 
    - Please provide honest and constructive feedback. This team feedback will be graded for completion.

- HW 03 **due Mon, Feb 24 at 11:59p**

---

### Today's Agenda

- Interactions 

- Log Transformations 


```{r echo = F}
library(tidyverse)
library(knitr)
library(broom)
library(Sleuth3) 
library(patchwork)
```
---

class: middle, center

## Interactions

---

### Checking for interactions in the EDA

```{r,echo=F, fig.height = 3.5}
case1202 <- case1202 %>%
  mutate(Z = if_else(Sex == "Female", "Group 1", "Group 2"))

p1 <- ggplot(data=case1202, aes(x=Senior,y=Bsal,color = Z)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  labs(x = "X1", 
       y = "Y", 
       color = "Z") + 
  theme(legend.position = "bottom")

p2 <- ggplot(data=case1202, aes(x=Exper,y=Bsal,color = Z)) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) + 
  labs(x = "X2", 
       y = "Y", 
       color = "Z") + 
  theme(legend.position = "bottom")

patchwork <- p1 + p2
patchwork + plot_annotation(tag_levels = 'A')
```

---

### Model with interactions 



---


class: middle, center

## Log Transformations 

---

## Respiratory Rate vs. Age

- A high respiratory rate can potentially indicate a respiratory infection in children. In order to determine what indicates a "high" rate, we first want to understand the relationship between a child's age and their respiratory rate. 

- The data contain the respiratory rate for 618 children ages 15 days to 3 years.

- **Variables**: 
    - <font class="vocab">`Age`</font>: age in months
    - <font class="vocab">`Rate`</font>: respiratory rate (breaths per minute)

---

## Rate vs. Age

```{r}
respiratory <- ex0824
ggplot(data=respiratory, aes(x=Age, y=Rate)) +
  geom_point() + 
  labs("Respiratory Rate vs. Age")
```

---

## Rate vs. Age

```{r echo=FALSE}
model <- lm(Rate ~ Age, data=respiratory)
kable(tidy(model, conf.int=TRUE),format="html", digits=3)
```

```{r echo=FALSE}
respiratory <- respiratory %>% 
  mutate(residuals = resid(model), predicted = predict(model))

ggplot(data=respiratory, aes(x=predicted, y=residuals)) +
  geom_point() + 
  geom_hline(yintercept=0,color="red")
```

---

class: middle, center

## Log transformations

---

### Need to transform $y$ 

- Typically, a "fan-shaped" residual plot indicates the need for a transformation of the response variable $y$
  + $\mathbf{\color{green}{\log(y)}}$: Easiest to interpret 

--


- When building a model: 
  + Choose a transformation and build the model on the transformed data
  + Reassess the residual plots
  + If the residuals plots did not sufficiently improve, try a new transformation!

---

### Log transformation on $y$

- Use when the residual plot shows "fan-shaped" pattern

- If we apply a log transformation to the response variable, we want to estimate the parameters for the model...
.alert[
$$\log(y) = \beta_0 + \beta_1 x$$
]

--

- We want to interpret the model in terms of $y$ not $\log(y)$, so we write all interpretations in terms of 

.alert[
$$y = \exp\{\beta_0 + \beta_1 x\} = \exp\{\beta_0\}\exp\{\beta_1x\}$$
]

---

### Mean and logs

Suppose we have a set of values

```{r}
x <- c(3, 5, 6, 8, 10, 14, 19)
```

--

Let's find the mean of the logged values of x, i.e. $\overline{\log(x)}$

```{r}
log_x <- log(x)
mean(log_x)
```

--

Let's find mean of x and then log the mean value, i.e. $\log(\bar{x})$

```{r}
xbar <- mean(x)
log(xbar)
```

---

### Median and logs

```{r}
x <- c(3, 5, 6, 8, 10, 14, 19)
```

--

Let's find the median of the logged values of x, i.e. $\text{Median}(\log(x))$

```{r}
log_x <- log(x)
median(log_x)
```

--

Let's find median of x and then log the mean value, i.e. $\log(\text{Median}(x))$

```{r}
median_x <- median(x)
log(median_x)
```


---

### Mean, Median, and log

```{r}
x <- c(3, 5, 6, 8, 10, 14, 19)
```

--

$$\overline{\log(x)} \neq \log(\bar{x})$$

```{r}
mean(log_x) == log(xbar)
```

--

$$\text{Median}(\log(x)) = \log(\text{Median}(x))$$
```{r}
median(log_x) == log(median_x)
```


---

### Mean and median of $\log(y)$

- Recall that $y = \beta_0 + \beta_1 x_i$ is the **mean** value of $y$ at the given value $x_i$. This doesn't hold when we log-transform $y$

--

- The mean of the logged values is **not** equal to the log of the mean value. Therefore at a given value of $x$

.alert[
$$\begin{aligned}\exp\{\text{Mean}(\log(y))\} \neq \text{Mean}(y) \\[5pt]
\Rightarrow \exp\{\beta_0 + \beta_1 x\} \neq \text{Mean}(y) \end{aligned}$$
]

---

### Mean and median of $\log(y)$

- However, the median of the logged values **is** equal to the log of the median value. Therefore,

.alert[
$$\exp\{\text{Median}(\log(y))\} = \text{Median}(y)$$
]

--

- If the distribution of $\log(y)$ is symmetric about the regression line, for a given value $x_i$,
.alert[
$$\text{Median}(\log(y)) = \text{Mean}(\log(y))$$
]
---

### Interpretation with log-transformed $y$

- Given the previous facts, if $\log(y) = \beta_0 + \beta_1 x$, then 
.alert[
$$\text{Median}(y) = \exp\{\beta_0\}\exp\{\beta_1 x\}$$
]

<br><br>


- <font class="vocab">Intercept:</font> When $x=0$, the median of $y$ is expected to be $\exp\{\beta_0\}$
<br>


- <font class="vocab">Slope: </font>For every one unit increase in $x$, the median of $y$ is expected to multiply by a factor of $\exp\{\beta_1\}$


---

### log(Rate) vs. Age

```{r}
respiratory <- respiratory %>% mutate(log_rate = log(Rate))
```

```{r echo=F}
ggplot(data=respiratory, aes(x=Age,y=log_rate)) + 
  geom_point() +
  labs(title="Log-Transformed Rate vs. Age")
```

---

### log(Rate) vs. Age

```{r}
log_model <- lm(log_rate ~ Age, data = respiratory)
```

```{r echo=F}
ggplot(data=respiratory, aes(x=predict(log_model), y=resid(log_model))) +
  geom_point() + 
  geom_hline(yintercept=0, color="red") +
  labs(x="Predicted", y="Residuals",
       title="Residuals vs. Predicted", 
       subtitle="log-transformed Rate")
```

---

### log(Rate) vs. Age

```{r echo = F}
kable(tidy(log_model, conf.int=TRUE),format="html", digits=3)
```
<br>

.question[
1. Write the model in terms of $\log(rate)$. 
2. Write the model in terms of $rate$. Interpret the slope and intercept.

]

---

### Confidence interval for $\beta_j$

- The confidence interval for the coefficient of $x$ describing its relationship with $\log(y)$ is

$$\hat{\beta}_j \pm t^* SE(\hat{\beta_j})$$

--

- The confidence interval for the coefficient of $x$ describing its relationship with $y$ is

.alert[
$$\exp\big\{\hat{\beta}_j \pm t^* SE(\hat{\beta_j})\big\}$$
]

---

### Coefficient of `Age`

```{r echo = F}
kable(tidy(log_model, conf.int=TRUE),format="html", digits=3)
```

.question[
Interpret the 95% confidence interval for the coefficient of `Age` in terms of *rate*.
]

---

### Log Transformation on $x$

.pull-left[
```{r,echo=F}
set.seed(1)
s <- ggplot2::diamonds %>% sample_n(100)
ggplot(data=s,aes(x=carat,y=log(price)))+
  geom_point(color="blue")+
  ggtitle("Scatterplot")+
  xlab("X")+
  ylab("Y") + 
  theme(plot.title = element_text(hjust = 0.5,size=20))
```
]

.pull-right[
```{r,echo=F}
mod2 <- lm(log(price) ~ carat, data=s)
s <- s %>% mutate(residuals = resid(mod2))
ggplot(data=s,aes(x=carat,y=residuals)) + 
geom_point()+
geom_hline(yintercept=0,color="red") +
  ggtitle("Residual Plot")+
  xlab("X")+
  ylab("residuals") + 
  theme(plot.title = element_text(hjust = 0.5,size=20))
```
]

- Try a transformation on $X$ if the scatterplot shows some curvature but the variance is constant for all values of $X$

---

### Model with Transformation on $x$

.alert[
$$y = \beta_0 + \beta_1 \log(x)$$
]
<br> 

--

- <font class="vocab">Intercept: </font> When $\log(x)=0$, $(x=1)$, $y$ is expected to be $\beta_0$ (i.e. the mean of $y$ is $\beta_0$)

--

- <font class="vocab">Slope: </font> When $x$ is multiplied by a factor of $\mathbf{C}$, $y$ is expected to change by $\boldsymbol{\beta_1}\mathbf{\log(C)}$ units, i.e. the mean of $y$ changes by $\boldsymbol{\beta_1}\mathbf{\log(C)}$
    - *Example*: when $x$ is multiplied by a factor of 2, $y$ is expected to change by $\boldsymbol{\beta_1}\mathbf{\log(2)}$ units

---

### Rate vs. log(Age)

```{r,echo=F}
ex0824 <- ex0824 %>% mutate(log.age = log(Age))
ggplot(data=ex0824,aes(x=log.age,y=Rate)) + 
  geom_point()  +
  ggtitle("Respiratory Rate vs. log(Age)") + 
  xlab("log(Age)")+
  ylab("Respiratory Rate")
```

---

### Rate vs. Age

```{r,echo=F}
mod3 <- lm(Rate ~ log.age,data=ex0824)
kable(tidy(mod3, conf.int=TRUE),format="html")
```
<br> 

.question[

1. Write the equation for the model of $y$ regressed on $\log(x)$. 

2. Interpret the intercept in the context of the problem. 

3. Interpret the slope in terms of how the mean respiratory rate changes when a child's age doubles. 

4. Suppose a doctor has a patient who is currently 3 years old. Will this model provide a reliable prediction of the child's respiratory rate when her age doubles? Why or why not?
]


---

class: middle 

See [Log Transformations in Linear Regression](https://github.com/sta210-sp20/supplemental-notes/blob/master/log-transformations.pdf) for more details about interpreting regression models with log-transformed variables.


