---
title: "Exam 01 Review"
author: "Prof. Maria Tackett"
date: "02.24.20"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3,
	message = FALSE,
	warning = FALSE, 
	dpi = 300
)
```

class: middle, center

### [Click for PDF of slides](10-exam-01-review.pdf)

---

### Announcements

- HW 03 **due TODAY at 11:59p**

- Lab 06 - Spatial regression due **Tuesday at 11:59p**

- <font color = "blue">Sign Up for DataFest!</font>

  `r emo::ji("spiral_calendar")` April 3 - 5
  
  `r emo::ji("round_pushpin")` Penn Pavilion
  
  `r emo::ji("link")` [stat.duke.edu/datafest](https://www.stat.duke.edu/datafest)


```{r,echo=F}
library(knitr)
library(broom)
library(tidyverse)
```

---

### Exam I Outline

- Multiple choice, true/fallse, short answer questions

- Permitted to bring one sheet of <u>handwritten notes</u> (front and back)
  + Must show your notes to me or TA when you turn in exam. 
  + Points deducted if notes are not handwritten

- Calculator **not** permitted on exam

- Please use blue or black ink pen and write carefully!

---

### Topics 

- Regression Basics
  + Interpreting coefficients
  + Inference for regression
  + Predicted values and the appropriate interval
  + Regression assumptions

- ANOVA to compare group means

---

### Topics 

- Indicator variables and interaction terms

- Transformations on response variable
  + Determining if transformation is required
  + Interpreting coefficients on log-transformed and original scales
  + Calculating predicted values on log-transformed and original scales

---

class: middle, center 

## Questions?

---

class: middle, center

## Practice 

---

### Diamonds analysis

- Sit with your lab group. 

- Write your responses to each question. 

- You can download the .Rmd file [here](http://www2.stat.duke.edu/courses/Spring20/sta210.001/slides/lec-slides/10-exam-01-review-exercises.Rmd) and open it in RStudio if you wish to type your responses directly in the R Markdown file.

- We will spend a few minutes on each section, then select a few groups to share their responses. 

---

class: middle, center


### [Part I: Analysis of Variance](http://www2.stat.duke.edu/courses/Spring20/sta210.001/slides/lec-slides/10-exam-01-review-exercises.html)



```{r echo = F}
library(countdown)

countdown(minutes = 10, seconds = 0, warn_when = 60, right = "33%", bottom = "20%")
```

---

class: middle, center

### [Part II: Original Model](http://www2.stat.duke.edu/courses/Spring20/sta210.001/slides/lec-slides/10-exam-01-review-exercises.html#part_ii:_original_model)

```{r echo = F}
library(countdown)

countdown(minutes = 10, seconds = 0, warn_when = 60, right = "33%", bottom = "20%")
```

---

class: middle, center

### [Part III: Model with Log-Transformed Response](http://www2.stat.duke.edu/courses/Spring20/sta210.001/slides/lec-slides/10-exam-01-review-exercises.html#part_iii:_model_with_log-transformed_response)

```{r echo = F}
library(countdown)

countdown(minutes = 10, seconds = 0, warn_when = 60, right = "33%", bottom = "20%")
```