---
title: "Exam 01 Review"
author: "Dr. Maria Tackett"
date: "02.20.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3,
	message = FALSE,
	warning = FALSE, 
	dpi = 300
)
```


## Announcements

- Lab 05 due today at 11:59p 

- Team Feedback 1 due Friday at 11:59p 
    - will get an email from Teammates around noon
    
- Exam OH for Friday's lab
    
- HW 04 due Mon, Feb 25 at 11:59p

---

```{r,echo=F}
library(knitr)
library(broom)
library(tidyverse)
```

---

## Exam I Outline

- Short answer questions

- Permitted to bring one sheet of handwritten notes (front and back)
  + Must turn in notes with exam 

- Calculator **not** permitted on exam

- Please use pencil!

---

## Topics 

--

- Regression Basics
  + Interpreting coefficients
  + Inference for regression
  + Predicted values and the appropriate interval
  + Regression assumptions


--

- ANOVA for regression & comparing means


---

 

## Topics 

- Transformations on the predictor and/or response variables
  + Determining if transformation is required
  + Interpreting coefficients on transformed and original scales
  + Calculating predicted values on transformed and original scales
  
  
--

- $R^2$ versus  Adjusted $R^2$

--

- Special predictors: indicator variables and interaction terms

---

## Topics 

- Interaction terms
--

- Nested F Test
--

- Influential points
  + Basic ideas - no calculations
--

- Multicollinearity 
  + Basic ideas - no calculations

---

class: middle, center 

## Questions?

---

## Diamonds Analysis 

- Copy the **Diamonds - Exam 01 Review** project on RStudio Cloud.