---
title: "Exam 01 Review"
author: "Dr. Maria Tackett"
date: "10.09.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3,
	message = FALSE,
	warning = FALSE, 
	dpi = 300
)
```

class: middle, center

### [Click for PDF of slides](13-exam-01-review.pdf)

---

## Announcements

- HW 04 **due Thursday at 11:59p**

- Thursday's Lab: Exam help hours
    
---

```{r,echo=F}
library(knitr)
library(broom)
library(tidyverse)
```

---

## Exam I Outline

- Short answer questions

- Permitted to bring one sheet of <u>handwritten notes</u> (front and back)
  + Must turn in notes with exam. Put your name at the top of your notes.
  + You will <u>not</u> be permitted to use your notes if they are not handwritten

- Calculator **not** permitted on exam

- Please use blue or black ink pen!

---

## Topics 

- Regression Basics
  + Interpreting coefficients
  + Inference for regression
  + Predicted values and the appropriate interval
  + Regression assumptions

- ANOVA to compare group means

- ANOVA for regression


---

## Topics 

- Transformations on the predictor or response variables
  + Determining if transformation is required
  + Interpreting coefficients on transformed and original scales
  + Calculating predicted values on transformed and original scales

- $R^2$ versus  Adjusted $R^2$

- Special predictors: indicator variables and interaction terms

---

## Topics 

- Interaction terms

- Nested F Test

- High leverage, outliers, and influential points
  + Basic ideas - no calculations

- Multicollinearity 
  + Basic ideas - no calculations

---

class: middle, center 

## Questions?

---

class: middle, center

## Practice 

---

## Cars analysis 

- Sit with your lab group. 

- Copy the **Exam 01 Review** project on RStudio Cloud to type your responses in the R Markdown file.

- You can also find the application exericse on the course schedule.
