---
title: "Logistic regression"
subtitle: "Exploratory data analysis & Drop-in-Deviance Test"
author: "Dr. Maria Tackett"
date: "03.20.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 4,
	fig.width = 7,
	message = FALSE,
	warning = FALSE
)
```


### Announcements

- HW 05 due Mon, March 25 at 11:59p

- Project proposal due Fri, March 29 at 11:59p

---

### Packages

```{r,echo=T}
library(tidyverse)
library(knitr)
library(broom)
library(NHANES)
library(pROC) #ROC curves
```

---

### NHANES Data

- [National Health and Nutrition Examination Survey](https://www.cdc.gov/nchs/nhanes/index.htm) is conducted by the National Center for Health Statistics (NCHS) 

- The goal is to *"assess the health and nutritional status of adults and children in the United States"*

- This survey includes an interview and a physical examination

---

### NHANES: Physical Activity & Sleep

- Does physical activity reduce the odds of having sleep trouble?

- We will analyze the following variables: 

  + <font class="vocab">`SleepTrouble`: </font>Participant has told doctor or other health professional they had trouble sleeping
  
  + <font class="vocab">`PhysActive`: </font>Participant does moderate to vigorous-intensity sports, fitness or recreational activities
  
  + <font class="vocab">`Age`: </font>Age at time of screening (in years). Participants 80 or older were recorded as 80.
  
  + <font class="vocab">`Education`: </font>8th Grade, 9-11th Grade, High School, Some College, College Grad
  
---

### NHANES: Physical Activity & Sleep

.small[
```{r echo=F}
set.seed(1234)
nhanes <- NHANES %>% 
  select(SleepTrouble, PhysActive, Age, Education) %>%
  drop_na() %>%
  filter(Age > 18) %>%
  sample_n(500)
nhanes %>% 
  slice(1:5)
```
]

---

class: middle, center

## Exploratory Data Analysis 

---

### Exploratory Data Analysis (EDA)

**Categorical predictors**: 
- Examine the percentage of $Y=1$ for each level (category)
- You can visualize using a stacked bar chart

**Quantitative predictors**:
- Order the predictor variable and divide into $g$ groups of equal size
- For each group, plot proportion of $Y=1$ versus the average value of the predictor
- See if there's a pattern (e.g. is there a quadratic pattern?)
- Use `binnedplot` function 

---

### EDA: Sleep trouble vs. Age

.small[
```{r}
# create 1/0 indicator for response variable to use binnedplot
nhanes <- nhanes %>% 
  mutate(sleep_tr_ind = if_else(SleepTrouble == "Yes",1,0))

arm::binnedplot(x = nhanes$Age, y = nhanes$sleep_tr_ind, 
                xlab = "Age", ylab = "proportion",
                main = "Sleep Trouble vs. Age", 
                col.int = "white" #remove intervals
                )
```
]

---

### Sleep trouble vs. Physical Activity 

```{r}
ggplot(data = nhanes, aes(x = PhysActive, fill = SleepTrouble)) +
  geom_bar(position = "fill") + 
  labs(y = "proportion", 
       title = "Sleep trouble vs. Physical activity")
```


---

### Interaction? 

- Plot Sleep trouble vs. Age for each level of Physical Activity

```{r echo=F, fig.height = 3}

phys <- nhanes %>% filter(PhysActive == "Yes")

arm::binnedplot(x = phys$Age, y = nhanes$sleep_tr_ind, 
                xlab = "Age",
                ylab = "proportion",
                main = "Phys Active: Sleep Trouble vs. Age", 
                col.int = "white" #remove intervals
                )

not_phys <- nhanes %>% filter(PhysActive == "No")

arm::binnedplot(x = not_phys$Age, y = nhanes$sleep_tr_ind, 
                xlab = "Age",
                ylab = "proportion",
                main = "Not Phys Active: Sleep Trouble vs. Age", 
                col.int = "white" #remove intervals
                )
```

---

class: middle, center

### Drop-in-deviance test

---

### Comparing Nested Models 

- Suppose there are two models: 
    - Model 1 includes predictors $x_1, \ldots, x_q$
    - Model 2 includes predictors $x_1, \ldots, x_q, x_{q+1}, \ldots, x_p$

- We want to test the hypotheses
$$\begin{aligned}&H_0: \beta_{q+1} = \dots = \beta_p = 0 \\
& H_a: \text{ at least 1 }\beta_j \text{ is not} 0\end{aligned}$$

--

- We used a *Nested F Test* to compare two nested models in linear regression 

- We will use the <font class="vocab3">drop-in-deviance test</font> in logistic regression 

---

### Deviance residual 

- The **deviance residual** is the a measure of how much the observed data differs from what is measured using the likelihood ratio 

- The deviance residual for the $i^{th}$ observation is 

.alert[
$$d_i = \text{sign}(Y_i - \hat{p}_i)\sqrt{2\bigg[Y_i \log\Big(\frac{Y_i}{p_i}\Big) + (1-Y_i)\log\Big(\frac{1-Y_i}{1-\hat{p}_i}\Big)\bigg]}$$
where $\text{sign}(Y_i - \hat{p}_i)$ is positive when $Y_1 = 1$ and negative when $Y_i = 0$.
]




---

### Drop-in-Deviance Test

- The **deviance statistic** for Model $k$ is $D_k = \sum\limits_{i=1}^n d_i^2$

--

- To test the hypotheses
$$\begin{aligned}&H_0: \beta_{q+1} = \dots = \beta_p = 0 \\
& H_a: \text{ at least 1 }\beta_j \text{ is not} 0\end{aligned}$$

--

the <font class="vocab"> drop-in-deviance statistic </font> is $D_1 - D_2$

--

- When the sample size is large, the drop-in-deviance statistic has an approximately Chi-squared distribution with degrees of freedom equal to the difference in number of predictor variables in Model 1 and Model 2

---

### Should we add `Education` to the model?

- Suppose 
    - Model 1 includes `Age`, `PhysActive`, `Age*PhysActive`
    - Model 2 includes `Age`, `PhysActive`, `Age*PhysActive`, `Education`

.small[
```{r}
model1 <- glm(SleepTrouble ~ Age + PhysActive + Age * PhysActive, 
              data = nhanes, family = binomial)
model2 <- glm(SleepTrouble ~ Age + PhysActive + Age * PhysActive + 
                Education, data = nhanes, family = binomial)
```
]

--

```{r}
# Deviances
(dev_model1 <- glance(model1)$deviance)
(dev_model2 <- glance(model2)$deviance)
```

---
### Should we add `Education` to the model?

```{r}
# Drop-in-deviance test statistic
(test_stat <- dev_model1 - dev_model2)
```

--

```{r}
# p-value
1 - pchisq(test_stat, 4)
```

---

### Should we add `Education` to the model?

- We can use the **`anova`** function to conduct this test 
    - Add **`test = "Chisq"`** to conduct the drop-in-deviance test

```{r}
anova(model1, model2, test = "Chisq")
```

---

## Framingham Study

- We will analyze data from a cardiovascular study on residents in Framingham, MA

- **Goal**: Predict whether or not a participant has a 10-year risk of future coronary heart disease

- Original data contains information from 4,000+ participants. We will use 500 for this analysis.


