---
title: "Logistic regression"
subtitle: "Model Predictions"
author: "Dr. Maria Tackett"
date: "10.28.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE
)
```

class: middle, center

### [Click for PDF of slides](17-logistic-pt2.pdf)
---

### Announcements

- Lab 07 **due Tue, Oct 29 at 11:59p**

- [Reading 09](https://www2.stat.duke.edu/courses/Fall19/sta210.001/reading/reading-09.html) for Wednesday 

- Project Proposal **due Wed, Oct 30 at 11:59p**

---

 

### Packages

```{r,echo=T}
library(tidyverse)
library(knitr)
library(broom)
library(pROC) #ROC curves
```

---

### Review

- $y$: binary response
  + 1: yes
  + 0: no
  
- $Mean(y) = p$

- $Var(y) = p(1-p)$

- **Odds of "yes"**: $\omega = \frac{p}{1-p}$

---

### Logistic Regression Model 

- Suppose $P(y_i=1|x_i) = p_i$ and $P(y_i=0|x_i) = 1-p_i$

- The <font class="vocab3">logistic regression model </font> is

$$\log\Big(\frac{p_i}{1-p_i}\Big) = \beta_0 + \beta_1 x_i$$
<br> 


- $\log\Big(\frac{p_i}{1-p_i}\Big)$ is called the <font class="vocab3">logit</font> function


---

### Interpreting the intercept: $\beta_0$

.alert[
$$\log\Big(\frac{p_i}{1-p_i}\Big) = \beta_0 + \beta_1 x_i$$
]

- Intercept: When $x=0$, odds of $y$ are $\exp\{\beta_0\}$

---

### Interpreting slope coefficient $\beta_1$

.alert[
$$\log\Big(\frac{p_i}{1-p_i}\Big) = \beta_0 + \beta_1 X_i$$
]

If $x$ is a <u>quantitative</u> predictor

- **As $x_i$ increases by 1 unit, we expect the odds of $y$ to multiply by a factor of $\exp\{\beta_1\}$**

If $x$ is a <u>categorical</u> predictor

- **The odds of $y$ for group $k$ are expected to be $\exp\{\beta_1\}$ times the odds of $y$ for the baseline group.**

---

### Inference for coefficients

- The standard error is the estimated standard deviation of the sampling distribution of $\hat{\beta}_1$

- We can calculate the $\color{blue}{C%}$ <font color="blue">confidence interval</font> based on the large-sample Normal approximations

- **CI for $\boldsymbol{\beta}_1$**: $$\hat{\beta}_1 \pm z^* SE(\hat{\beta}_1)$$

.alert[
**CI for $\exp\{\boldsymbol{\beta}_1\}$**: $$\exp\{\hat{\beta}_1 \pm z^* SE(\hat{\beta}_1)\}$$
  ]

---

### Risk of coronary heart disease 

This data is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. The goal is to predict whether a patient has a 10-year risk of future coronary heart disease.

**Response**: 

.vocab[`TenYearCHD`]: 
- 0 = Patient doesn’t have 10-year risk of future coronary heart disease
- 1 = Patient has 10-year risk of future coronary heart disease

**Predictor**: 

- .vocab[`age`]: Age at exam time.
- .vocab[`currentSmoker`]: 0 = nonsmoker; 1 = smoker
- .vocab[`totChol`]: total cholesterol (mg/dL)


```{r echo = F}
heart_data <- read_csv("data/framingham.csv") %>%
  drop_na() %>%
  mutate(currentSmoker = as.factor(currentSmoker), 
         TenYearCHD = as.factor(TenYearCHD), 
         ageCent = age - mean(age), 
         totCholCent = totChol - mean(totChol))
```


---

### Modeling risk of coronary heart disease

```{r echo = F}
risk_m <- glm(TenYearCHD ~ ageCent + currentSmoker + totCholCent, data = heart_data, 
         family = binomial)
tidy(risk_m, conf.int = TRUE) %>% kable(format = "markdown", digits = 3)
```

.question[
1. Interpret `age` in terms of the odds of being at risk for coronary heart disease.
2. Interpret `currentSmoker1` in terms of the odds of being at risk for coronary heart disease.
]

---

class: middle, center

### Predictions & Model Fit

---

### Predictions

- We are often interested in predicting if a given observation will have a "yes" response 

- To do so, we will use the logistic regression model to predict the probability of a "yes" response for the given observation. If we have one predictor variable, then...

$$p_i = \frac{\exp(\beta_0 + \beta_1 x_i)}{1 + \exp(\beta_0 + \beta_1 x_i)}$$

- We will use the predicted probabilities to classify the observation as having a "yes" or "no" response

---

### Is the patient at risk for coronary heart disease?

- Suppose a patient comes in who is 60 years old, does not currently smoke, and has a total cholesterol of 263 mg/dL

--

- Predicted log-odds that this person is at risk for coronary heart disease: 

$$\begin{aligned}\log\bigg(\frac{\hat{p}_i}{1-\hat{p}_i}\bigg) = &-2.111 + 0.081\times(60 - 49.552) - 0.447 \times 0 \\
&+ 0.002 \times (263 - 236.848)\approx -1.212\end{aligned}$$

--

- The probability this passenger thinks reclining the seat is rude: 

$$\hat{p}_i = \frac{\exp\{ -1.212\}}{1 +  \exp\{-1.212\}} = 0.229$$

---

### Predictions in R

```{r}
x0 <- data_frame(ageCent = (60 - 49.552), totCholCent = (263 - 236.848), 
currentSmoker = as.factor(0))
```

- **Predicted log-odds**

```{r}
predict(risk_m, x0) 
```

- **Predicted probabilities**

```{r}
predict(risk_m, x0, type = "response") 
```

---

### Is the patient at risk for coronary heart disease?


```{r}
predict(risk_m, x0, type = "response") 
```

The probability the patient is at risk for coronary heart disease is `r round(predict(risk_m, x0, type = "response"), 3)`.

.question[
Based on this probability, would you consider the patient at risk for coronary heart disease? Why or why not?
]

---


### Confusion Matrix

- We can use the estimated probabilities to predict outcomes 

- *Ex.*: Establish a threshold such that $y=1$ if predicted probability is greater than the threshold ($y=0$ otherwise)

- Determine how many observations were classified correctly and incorrectly and put the results in a $2 \times 2$ table
  + This table is the <font class="vocab3">confusion matrix</font>

- If the proportion of misclassifications is high, then we might conclude the model doesn't fit the data well

---

### Confusion Matrix

Suppose we use 0.3 as the threshold to classify responses

```{r}
threshold <- 0.3
risk_m_aug <- augment(risk_m, type.predict = "response")
```

```{r}
risk_m_aug %>%
  mutate(risk_predict = if_else(.fitted > threshold, "Yes", "No")) %>%
  group_by(TenYearCHD, risk_predict) %>%
  summarise(n = n()) %>%
  spread(TenYearCHD, n) %>%
  kable(format="markdown")
```

---

### Confusion matrix

```{r echo=F}
risk_m_aug %>%
  mutate(risk_predict = if_else(.fitted > threshold, "Yes", "No")) %>%
  group_by(TenYearCHD, risk_predict) %>%
  summarise(n = n()) %>%
  spread(TenYearCHD, n) %>%
  kable(format="markdown")
```
<br><br>

.question[ 
What proportion of observations were misclassified?]

---

### Sensitivity & Specificity

- <font class="vocab3">Sensitivity: </font>Proportion of observations with $y=1$ that have predicted probability above a specified threshold
  + Called true positive rate

- <font class="vocab3">Specificity: </font>Proportion of observations with $y=0$ that have predicted probability below a specified threshold
  + (1 - specificity) called false positive rate

- What we want: 
  + High sensitivity
  + Low values of 1-specificity

---

class: regular 

### ROC Curve

- <font class="vocab3">Receive Operating Characteristic (ROC) curve </font>: 
  + *x-axis*: $1 - \text{ specificity}$
  + *y-axis*: $\text{ Sensitivity}$ 
  
- Evaluated with a lot of different values for the threshold

- Logistic model fits well if the area under the curve (AUC) is close to 1

- ROC in R
    - Use the <font class="vocab">`roc`</font> function in the `pROC` to calculate AUC
    - Use <font class="vocab">`geom_roc`</font> layer in ggplot to plot the ROC curve

---

### Visualize ROC curve

```{r}
library(plotROC) #extension of ggplot2
roc_curve <- ggplot(risk_m_aug, aes(d = as.numeric(TenYearCHD) -1, m = .fitted)) +
  geom_roc(n.cuts = 5, labelround = 3) + 
  geom_abline(intercept = 0) + 
  labs(x = "False Positive Rate", y = "True Positive Rate") 
roc_curve
```


---

### Area under curve

```{r echo = F}
roc_curve
```

```{r}
calc_auc(roc_curve)$AUC
```

---

### Application Exercise

Copy the **Logistic Regression** project on RStudio Cloud

