---
title: "Logistic regression"
subtitle: "Model fit & Exploratory data analysis"
author: "Dr. Maria Tackett"
date: "10.30.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 4,
	fig.width = 7,
	message = FALSE,
	warning = FALSE
)
```

class: middle, center

### [Click for PDF of slides](18-logistic-pt3.pdf)
---

### Announcements

- [Reading 10](https://www2.stat.duke.edu/courses/Fall19/sta210.001/reading/reading-10.html) for Monday

- Project Proposal **due TODAY at 11:59p**

- [Electronic Undergraduate Research Conference](https://www2.stat.duke.edu/courses/Fall19/sta199.001/slides/lec-slides/eusr_ad.pdf)

- HW 05 **due Wed, Nov 6 at 11:59p**

---

### Packages

```{r,echo=T}
library(tidyverse)
library(knitr)
library(broom)
library(pROC) #ROC curves
```

---

### Risk of coronary heart disease 

This data is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts. The goal is to predict whether a patient has a 10-year risk of future coronary heart disease.

**Response**: 

.vocab[`TenYearCHD`]: 
- 0 = Patient doesn’t have 10-year risk of future coronary heart disease
- 1 = Patient has 10-year risk of future coronary heart disease

**Predictor**: 

- .vocab[`age`]: Age at exam time.
- .vocab[`currentSmoker`]: 0 = nonsmoker; 1 = smoker
- .vocab[`totChol`]: total cholesterol (mg/dL)


```{r echo = F}
heart_data <- read_csv("data/framingham.csv") %>%
  drop_na() %>%
  mutate(currentSmoker = as.factor(currentSmoker), 
         ageCent = age - mean(age), 
         TenYearCHD = as.factor(TenYearCHD),
         totCholCent = totChol - mean(totChol))
```

---

### Logistic Regression Model 

- Suppose $P(Y_i=1|X_i) = p_i$ and $P(Y_i=0|X_i) = 1-p_i$

- The <font class="vocab3">logistic regression model </font> is

$$\log\Big(\frac{p_i}{1-p_i}\Big) = \beta_0 + \beta_1 X_i$$
<br> 


- $\log\Big(\frac{p_i}{1-p_i}\Big)$ is called the <font class="vocab3">logit</font> function

---

### Modeling risk of coronary heart disease

```{r echo = F}
risk_m <- glm(TenYearCHD ~ ageCent + currentSmoker + totCholCent, data = heart_data, 
         family = binomial)
tidy(risk_m, conf.int = TRUE) %>% kable(format = "markdown", digits = 3)
```

---

### Logistic Regression Model 

$$\log\Big(\frac{p_i}{1-p_i}\Big) = \beta_0 + \beta_1 X_i$$
<br> 
<br> 

- We can calculate $p_i$ by solving the logit equation: 

$$p_i = \frac{\exp\{\beta_0 + \beta_1 X_i\}}{1 + \exp\{\beta_0 + \beta_1 X_i\}}$$

---

### ROC Curve

- .vocab[Receiver Operating Characteristic (ROC) curve]: 
  + *X-axis*: $1 - \text{ specificity}$
  + *Y-axis*: $\text{ Sensitivity}$ 
  
- Evaluated with a lot of different values for the threshold

- Logistic model fits well if the area under the curve (AUC) is close to 1

- ROC in R
    - Use the <font class="vocab">`roc`</font> function in the `pROC` to calculate AUC
    - Use <font class="vocab">`geom_roc`</font> layer in ggplot to plot the ROC curve

---

### ROC Curve

.small[
```{r echo = F}
risk_m_aug <- augment(risk_m, type.predict = "response", type.residuals = "response")
```

```{r echo = F}
library(plotROC) #extension of ggplot2
roc_curve <- ggplot(risk_m_aug, aes(d = as.numeric(TenYearCHD) - 1, m = .fitted)) +
  geom_roc(n.cuts = 5, labelround = 3) + 
  geom_abline(intercept = 0) + 
  labs(x = "False Positive Rate", y = "True Positive Rate") 
roc_curve
calc_auc(roc_curve)$AUC
```

---

### Not useful: Raw residuals vs. predicted

- Include **`type.residuals = "response"`** in the `augment` function to get the raw residuals. 
$$e_i = Y_i - \hat{p}_i$$ 

```{r,echo=F}
#plot raw residuals vs. predicted
ggplot(data=risk_m_aug, aes(x=.fitted,y=.resid)) + 
  geom_point() + 
  geom_hline(yintercept=0,color="red")+
  labs(title="Raw Residuals vs. Predicted")
```

---

### Binned Residuals

- It is not useful to plot the raw residuals, so we will examine binned residual plots 

**When examining binned residuals**
    
- Look for patterns

- Nonlinear trend may be indication that squared term or log transformation of predictor variable required

- If bins have average residuals with large magnitude
  + Look at averages of other predictor variables across bins
  + Interaction may be required if large magnitude residuals correspond to certain combinations of predictor variables
  
---

### Binned plot vs. predicted values

- Use the <font class="vocab">`binnedplot`</font> function in the **arm** package. 
    - *Tip: Don't load the **arm** package to avoid conflicts with tidyverse*

.small[
```{r}
arm::binnedplot(x = risk_m_aug$.fitted, y = risk_m_aug$.resid,
                xlab = "Predicted Probabilities", 
                main = "Binned Residual vs. Predicted Values", 
                col.int = FALSE)
```
]

---

### Making binned residual plot

- Calculate raw residuals

- Order observations either by the values of the predicted probabilities (or by numeric predictor variable)

- Use the ordered data to create *g* bins of approximately equal size. Default value: $g = \sqrt{n}$

- Calculate average residual value in each bin

- Plot average residuals vs. average predicted probability (or average predictor value)

---

###  Residuals vs. `Age`

Make binned plot with predictor on $x$ axis

```{r}
arm::binnedplot(x = risk_m_aug$ageCent, 
                y = risk_m_aug$.resid, 
                col.int = FALSE,
                xlab = "Age (Mean-Centered)", 
                main = "Binned Residual vs. Age")
```

---

### Residuals vs. `totChol`


```{r}
arm::binnedplot(x = risk_m_aug$totCholCent, 
                y = risk_m_aug$.resid, 
                col.int = FALSE,
                xlab = "Total Cholesterol (Mean-Centered)", 
                main = "Binned Residual vs. Total Cholesterol")
```

---

### Residuals vs. categorical predictors 

- Calculate average residual for each level of the predictor
    - Are all means close to 0? If not, there is a problem with model fit.
    
```{r}
risk_m_aug %>%
  group_by(currentSmoker) %>%
  summarise(mean_resid = mean(.resid))
```

---

### Residuals 

Let's look at the binned residuals versus `AgeCent` separately for those who currently smoke and those who do not


```{r, echo=F, fig.height = 3}
 smoker <- risk_m_aug %>% filter(currentSmoker == 1)
arm::binnedplot(x=smoker$ageCent,y=smoker$.resid,xlab="AgeCent", main = "Current Smokers: Binned Residuals vs. Age", col.int = FALSE)

nonsmoker <- risk_m_aug %>% filter(currentSmoker == 0)
arm::binnedplot(x=nonsmoker$ageCent,y=nonsmoker$.resid,xlab="AgeCent", main = "Current Non-Smokers: Binned Residuals vs. Age", col.int = FALSE)
```

---

### Model with interaction term

```{r,echo=F}
risk_m_int <- glm(as.factor(TenYearCHD) ~ ageCent + totCholCent + currentSmoker + ageCent*currentSmoker,
                       data = heart_data, family = binomial)
kable(tidy(risk_m_int, exponentiate = FALSE, conf.int = TRUE), 
      format = "markdown", digits = 3)
```

.question[
1. What is the effect of age on the odds of being at risk of heart disease for smokers?  
2. What is the effect of age on the odds of being at risk of heart disease for non-smokers?  
3. Is the effect of age on being at risk of heart disease significantly different for the two groups?
]

---

### Binned residuals

```{r echo = F}
risk_m_int_aug <- augment(risk_m_int, type.predict = "response", 
                     type.residuals = "response")

arm::binnedplot(x=risk_m_int_aug$.fitted,y=risk_m_int_aug$.resid,xlab="Predicted Probabilities", main = "Binned Residuals vs. Predicted Probability", col.int = FALSE)
```

---

### Residuals vs. quantitative predictor

```{r}
arm::binnedplot(x=risk_m_int_aug$ageCent,y=risk_m_int_aug$.resid,xlab="Age (Mean-Centered)", main = "Binned Residuals vs. Age", col.int = FALSE)
```

---

### Residuals vs. categorical predictor

```{r echo=F}
risk_m_int_aug %>%
  group_by(currentSmoker) %>%
  summarise(mean_resid = mean(.resid))
```

---

### Binned Residuals vs. Age: Smokers vs. Non-Smokers

```{r, echo=F, fig.height = 3}
smoker <- risk_m_int_aug %>% filter(currentSmoker == 1)
arm::binnedplot(x=smoker$ageCent,y=smoker$.resid,xlab="Age (Mean-Centered)", main = "Smokers: Binned Residuals vs. Age", col.int = FALSE)

nonsmoker <- risk_m_int_aug %>% filter(currentSmoker == 0)
arm::binnedplot(x=nonsmoker$ageCent,y=nonsmoker$.resid,xlab="Age (Mean-Centered)", main = "Non-Smokers: Binned Residuals vs. Age", col.int = FALSE)
```

---

### Binned Residuals vs. total cholesterol

```{r echo = F}
arm::binnedplot(x=risk_m_int_aug$totCholCent,y=risk_m_int_aug$.resid,xlab="Total Cholesterol (Mean-Centered)", main = "Binned Residuals vs. Total Cholesterol", col.int = FALSE)
```

---

class: middle, center

## Exploratory Data Analysis 

---

### Exploratory Data Analysis 

**Categorical predictors**: 
- Examine the percentage of $y=1$ for each level (category)
- You can visualize using a stacked bar chart

**Quantitative predictors**:
- Use side-by-side boxplots to examine the distribution of the predictor for each level of the response
- Use binned plots to examine how the probability of $y=1$ changes as the predictor increases

---

### EDA: `TenYearCHD` vs. `Age`

.pull-left[
```{r echo = F}
# create 1/0 indicator for response variable to use binnedplot
ggplot(data = heart_data, aes(x = TenYearCHD, y = age)) +
  geom_boxplot() + 
  labs(x = "At Risk for Heart Disease", y = "Age", title = "Age vs. Risk of Heart Disease")
```
]

.pull-right[
```{r echo = F}
arm::binnedplot(x = heart_data$age, y = as.numeric(heart_data$TenYearCHD)-1, 
                xlab = "Age",
                ylab = "Proportion at Risk for Heart Disease",
                main = "Risk of Heart Disease vs. Age", 
                col.int = FALSE #remove intervals
                )
```
]


---

### EDA: `TenYearCHD` vs. `totChol`

.pull-left[
```{r echo = F}
# create 1/0 indicator for response variable to use binnedplot
ggplot(data = heart_data, aes(x = TenYearCHD, y = totChol)) +
  geom_boxplot() + 
  labs(x = "At Risk for Heart Disease", y = "Total Cholesterol (mg/dL)", title = "Total Cholesterol vs. Risk of Heart Disease")
```
]

.pull-right[
```{r echo = F}
arm::binnedplot(x = heart_data$totChol, y = as.numeric(heart_data$TenYearCHD)-1, 
                xlab = "Total Cholesterol (mg/dL)",
                ylab = "Proportion at Risk for Heart Disease",
                main = "Risk of Heart Disease vs. Total Cholesterol", 
                col.int = FALSE #remove intervals
                )
```
]
---

### `TenYearCHD` vs. `currentSmoker`

```{r}
ggplot(data = heart_data, aes(x = currentSmoker, fill = as.factor(TenYearCHD))) +
  geom_bar(position = "fill") + 
  labs(y = "Proportion", 
       fill = "At Risk of Heart Disease",
       title = "Risk of Heart Disease vs. Smoking Status")
```
---

class: middle, center

### Drop-in-deviance test

---

### Comparing Nested Models 

- Suppose there are two models: 
    - Model 1 includes predictors $x_1, \ldots, x_q$
    - Model 2 includes predictors $x_1, \ldots, x_q, x_{q+1}, \ldots, x_p$

- We want to test the hypotheses
$$\begin{aligned}&H_0: \beta_{q+1} = \dots = \beta_p = 0 \\
& H_a: \text{ at least 1 }\beta_j \text{ is not} 0\end{aligned}$$

--

- We used a *Nested F Test* to compare two nested models in linear regression 

- We will use the <font class="vocab3">drop-in-deviance test</font> in logistic regression 

---

### Deviance residual 

- The **deviance residual** is the a measure of how much the observed data differs from what is measured using the likelihood ratio 

- The deviance residual for the $i^{th}$ observation is 

.alert[
$$d_i = \text{sign}(Y_i - \hat{p}_i)\sqrt{2\bigg[Y_i \log\Big(\frac{Y_i}{p_i}\Big) + (1-Y_i)\log\Big(\frac{1-Y_i}{1-\hat{p}_i}\Big)\bigg]}$$
where $\text{sign}(Y_i - \hat{p}_i)$ is positive when $Y_1 = 1$ and negative when $Y_i = 0$.
]


---

### Drop-in-Deviance Test

- The **deviance statistic** for Model $k$ is $D_k = \sum\limits_{i=1}^n d_i^2$

--

- To test the hypotheses
$$\begin{aligned}&H_0: \beta_{q+1} = \dots = \beta_p = 0 \\
& H_a: \text{ at least 1 }\beta_j \text{ is not} 0\end{aligned}$$

--

the <font class="vocab"> drop-in-deviance statistic </font> is $D_1 - D_2$

--

- When the sample size is large, the drop-in-deviance statistic has an approximately Chi-squared distribution with degrees of freedom equal to the difference in number of predictor variables in Model 1 and Model 2

---

### Should we add `Education` to the model?

- Suppose 
    - Model 1 includes `AgeCent`, `currentSmoker`, `totCholCent`
    - Model 2 includes `AgeCent`, `currentSmoker`, `totCholCent`, `education` (categorical)

.small[
```{r}
model1 <- glm(TenYearCHD ~ ageCent + currentSmoker + totChol, 
              data = heart_data, family = binomial)
model2 <- glm(TenYearCHD ~ ageCent + currentSmoker + totChol + 
                as.factor(education), 
              data = heart_data, family = binomial)
```
]

--

```{r}
# Deviances
(dev_model1 <- glance(model1)$deviance)
(dev_model2 <- glance(model2)$deviance)
```

---
### Should we add `education` to the model?

```{r}
# Drop-in-deviance test statistic
(test_stat <- dev_model1 - dev_model2)
```

--

```{r}
# p-value
1 - pchisq(test_stat, 3) #3 = number of new model terms in model2
```

---

### Should we add `Education` to the model?

- We can use the **`anova`** function to conduct this test 
    - Add **`test = "Chisq"`** to conduct the drop-in-deviance test

```{r}
anova(model1, model2, test = "Chisq")
```

---

## Appex: Framingham Study

- We will analyze data from a cardiovascular study on residents in Framingham, MA

- **Goal**: Predict whether or not a participant has a 10-year risk of future coronary heart disease

- Original data contains information from 4,000+ participants. We will use 500 for this analysis.





