---
title: "Model Validation"
author: "Dr. Maria Tackett"
date: "04.01.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 4,
	fig.width = 7,
	message = FALSE,
	warning = FALSE
)
```


### Announcements

- Lab 08 due Wed at 11:59p

- HW 06 due Mon, Apr 8


```{r, echo = F}
library(tidyverse)
library(knitr)
library(broom)
```

---

class: middle, center

### Multinomial Logistic Regression Example
### *Sesame Street*
---

### Example: *Sesame Street*

- We will analyze data from an experiment to test the effectiveness of the children's program *Sesame Street*. 

- As part of the experiment, children were assigned to one of two groups: those who were encouraged to watch the program and those who were not

- We want to understand what effect the encouragement had on the frequency of viewing after adjusting for other factors

---

### Response Variable

- <font class="vocab">`viewcat`</font>
  + 1: rarely watched show
  + 2: once or twice a week
  + 3: three to five times a week
  + 4: watched show on average more than five times a week
  
---

### Predictor Variables

- <font class="vocab">`age`:</font> child's age in months
- <font class="vocab">`prenumb`: </font>score on numbers pretest (0 to 54)
- <font class="vocab">`prelet`: </font>score on letters pretest (0 to 58)
- <font class="vocab">`viewenc`:</font> 1: encouraged to watch, 2: not encouraged
- <font class="vocab">`site:`</font>
  + 1: three to five year old from disadvantaged inner city area
  + 2: four year old from advantaged suburban area
  + 3: from advantaged rural area
  + 4: from disadvantaged rural area
  + 5: from Spanish speaking home

.footnote[[Full data description](http://www2.stat.duke.edu/~jerry/sta210/sesamelab.html)]

---

class: middle, center

[*Sesame Street* analysis](https://www2.stat.duke.edu/courses/Spring19/sta210.001/appex/18-multinom-logistic.html)


