---
title: "Model Validation"
author: "Dr. Maria Tackett"
date: "04.01.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 4,
	fig.width = 7,
	message = FALSE,
	warning = FALSE
)
```


### Announcements

- Lab 08 due Wed at 11:59p

- Team feedback #2 due Wed at 11:59p

- HW 06 due Mon, Apr 8


```{r, echo = F}
library(tidyverse)
library(knitr)
library(broom)
library(modelr)
```

---

class: middle, center

### Multinomial Logistic Regression Example
### *Sesame Street*
---

### Example: *Sesame Street*

- We will analyze data from an experiment to test the effectiveness of the children's program *Sesame Street*. 

- As part of the experiment, children were assigned to one of two groups: those who were encouraged to watch the program and those who were not

- We want to understand what effect the encouragement had on the frequency of viewing after adjusting for other factors

---

### Response Variable

- <font class="vocab">`viewcat`</font>
  + 1: rarely watched show
  + 2: once or twice a week
  + 3: three to five times a week
  + 4: watched show on average more than five times a week
  
---

### Predictor Variables

- <font class="vocab">`age`:</font> child's age in months
- <font class="vocab">`prenumb`: </font>score on numbers pretest (0 to 54)
- <font class="vocab">`prelet`: </font>score on letters pretest (0 to 58)
- <font class="vocab">`viewenc`:</font> 1: encouraged to watch, 2: not encouraged
- <font class="vocab">`site:`</font>
  + 1: three to five year old from disadvantaged inner city area
  + 2: four year old from advantaged suburban area
  + 3: from advantaged rural area
  + 4: from disadvantaged rural area
  + 5: from Spanish speaking home

.footnote[[Full data description](http://www2.stat.duke.edu/~jerry/sta210/sesamelab.html)]

---

class: middle, center

## [Sesame Street analysis](https://www2.stat.duke.edu/courses/Spring19/sta210.001/appex/18-multinom-logistic.html)

---

class: middle, center

## Model Validation 

---

## Model Validation 

- <font class="vocab">Goal: </font> Want to find set of variables that give the lowest test (not training) error
    - Want a model that is generalizable, i.e. can be used to make predictions for new observations

- If we have a large data set, we can achieve this goal by randomly splitting the data into training and test (validation) sets

- Use the training set to fit a model, then use the fitted model to predict the responses for the observations in the test set 

- Assess the error when applied to the test set and choose the model with the lowest error
    
---

## Error

- **Quantitative response**: use Mean Square Error

.alert[
$$MSE = \frac{1}{n}\sum\limits_{i=1}^{n}(y_i - \hat{y}_i)^2$$
]

--

- **Categorical response**: use misclassification rate

.alert[
$$\text{Misclassification Rate} = \frac{1}{n}\sum\limits_{i=1}^{n}Err_{i} = \frac{1}{n}\sum\limits_{i=1}^{n} I(y_i \neq \hat{y}_i)$$
]

---

### Training and test set

- There is no set split for the training and test sets. Common splits are
    - 50% training; 50% test
    - 80% training; 20% test
 
- Assigning observations to training and test sets: 
    - Random assignment
    - Pick a certain time point to make split, if data is collected over time. Generally use earlier data in training and later data in test.
    - Use other relevant characteristic to make split 

---

## Cautions 

- Be sure the training set is large enough to build a reliable model    
    - The number of observations should be at least 10 times larger than the number of predictors 
    
- Standard errors for model coefficients fit using training data are larger than standard errors if entire dataset was used
    - If the training set is reasonably large, then the difference in standard errors is small
    
- The test error is highly variable depending on which observations are in the test set

---

## k-fold Cross Validation

- There are numerous validation methods that address the variability in testing error; we will focus on <font class="vocab">k-fold cross validation</font>
    - More in-depth discussion of model validation in STA 325

--

- **k-fold Cross Validation**
    + Randomly split the data into $k$ folds (typically 5 or 10)
    + Use $k-1$ folds to fit a model (this is the training data)
    + Assess how well model predicts on remaining fold (this is the test data) 
    + Repeat $k$ using a different fold as the test set each time

--

- Calculate estimated testing error by average the $k$ different error rates

--

- Once the variables for the final model have been selected, use the entire dataset to estimate coefficients for final model

---

### 5-fold Cross Validation in R

- Split data into 5 folds. Don't forget to set a seed!

```{r eval = F}
library(modelr)
set.seed(04012019)
mydata_cv <- crossv_kfold(my.data, 5)
```


- Fit model on each training set

```{r eval = F}
models <- map(mydata_cv$train, ~ lm(Y ~ X1 + ... + XP, data = .))
```

- Calculate MSE on each test set

```{r eval = F}
# map2_dbl in purrr package that's loaded with tidyverse
test_mse <- map2_dbl(models, mydata_cv$test, mse)
```

---

## Example: Advertising

We want to use spending on TV, radio, and newspaper advertising ($thousands) to predict total sales ($millions). The data contains the advertising and sales for 200 markets.

```{r}
advertising <- read_csv("data/advertising.csv")
glimpse(advertising)
```

We'll start by looking at the 5-fold cross validation results for the model using the predictors `radio` and `newspaper` 
---

### Adveritising: Split into 5 folds

.small[
```{r}
set.seed(04012019)
ad_cv <- crossv_kfold(advertising, 5)
ad_cv
```
]

---

### Advertising: Fit models

- Fit model on each training set

.small[
```{r}
models <- map(ad_cv$train, 
              ~ lm(sales ~ radio + newspaper, data = .))
models
```
]

---

### Advertising: Test error

.small[
```{r}
test_mse <- map2_dbl(models, ad_cv$test, mse)
test_mse
(error_model1 <- mean(test_mse))
```
]

---

## Advertising 

We will look at the cross-validation results for the model that includes `radio`, `newspaper`, and `tv` as predictors 

.small[
```{r}
models <- map(ad_cv$train, 
              ~ lm(sales ~ radio + newspaper + tv, data = .))
```
]

.small[
```{r}
test_mse <- map2_dbl(models, ad_cv$test, mse)
test_mse
(error_model2 <- mean(test_mse))
```
]

---

## Comparing Models

- The estimated testing error for 
    - Model 1: `radio` and `newspaper`: `r error_model1`
    - Model 2: `radio`, `newspaper`, `tv`: `r error_model2`
    
- Model 2 performs better than Model 1 when predicting the sales for new markets.



