---
title: "Modeling Longitudinal Data"
author: "Dr. Maria Tackett"
date: "12.04.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 4,
	fig.width = 7,
	message = FALSE,
	warning = FALSE
)
```


class: middle, center

### [Click for PDF of slides](26-long-data.pdf)

---

### Announcements


- Project write up and presentation **due Dec 10 at 11:59p**

- Project presentations on Dec 11
    - Lab 01L: 9a - 10:30a
    - Lab 02L: 10:30a - 12p

- Exam 2 extra credit:
  + 90% response rate on course eval: +1 pt on Exam 02 grades

- Thursday's lab: Project office hours 

- Office hours next week: Tuesday, 2p - 3:30p in Old Chem 118B


```{r,echo=F}
library(tidyverse)
library(cowplot)
library(tidyr)
library(knitr)
library(broom)
```

---

### US college graduation rates

**What factors most effect graduation rates at US colleges?**

<font class="vocab">Response variable: </font>

- **`rate`**: graduation rate, i.e. number of degrees awarded per 100 students enrolled

<font class = "vocab">Predictor variables: </font>

- **`year02`**: number of years since 2002
- **`faculty`**: mean number of full-time faculty in 2002 - 2009
- **`tuition`**: mean yearly tuition between 2002 and 2009

```{r}
college <- read_csv("data/colleges.csv") %>%
  filter(rate < 100) %>%
  mutate(year02 = year - 2002)
```

```{r echo = F}
college <- college %>%
  select(instname, year02, faculty, tuition, rate) 
```

---

### `college` data
 
.small[
```{r echo = F}
college %>% 
  filter(instname %in% c("Duke University", "University of North Carolina at Chapel Hill")) %>%
  mutate(instname = if_else(instname == "University of North Carolina at Chapel Hill", "UNC", instname))
```
]
---

### What makes this model different?

- **Goals**:  
    - Understand how the number of faculty members and tuition affects a college's graduation rate
    - How the graduation rate has changed over time

- There are multiple observations for each college (so multiple regression not appropriate)

- The are only a few time points and there's data on multiple colleges (so time series model not appropriate)

- We will use a <font class="vocab">multilevel model</font> to model the relationship between `faculty`, `tuition` and `rate`.


---

### Multilevel Model

We will fit a two-level model that includes the following model components: 

- **Level One**: include time  and any other predictors that change within a college over the time period in the data (`year02`)
    - The effects in this component are <font class = "vocab">random effects</font>
    - Typically not interested in drawing inferences about specific levels of these variables

- **Level Two**: includes predictors that differ between colleges but that remain the same within a college over the time period in the data (`faculty` and `tuition`)
    - The effects in this component are <font class = "vocab">fixed effects</font>
    - Typically the effects we wish to drawing inferences about
---

### Modeling Approach 

**Approach:** Start with simple, preliminary models to establish a baseline that can be used to evaluate more complex models. Work toward the final model by adding predictors and checking model assumptions at each step. We can take the following steps:

1. Exploratory data analysis 

2. Fit unconditional means model - model with no predictors

3. Fit unconditional growth model - add time

4. Fit "final" model with time and predictors

---

### 1. Exploratory Data Analysis

- Given the longitudinal structure of the data, we have observations at different time points for each college in the data set.

- When we do EDA, in addition to an univariate analysis of each variable, we want to look at the following: 
    - **within college**: changes over time within a school
    - **between college**: effects of school-specific predictors (e.g. `faculty`)

---

### 1. EDA: Univariate analysis

```{r echo = F, fig.height = 7}
p1 <- ggplot(data = college, aes(x = rate)) +
  geom_histogram() + 
  labs(title = "Graduation Rate")

p2 <- ggplot(data = college, aes(x = faculty)) +
  geom_histogram() + 
  labs(title = "Faculty")

p3 <- ggplot(data = college, aes(x = tuition)) +
  geom_histogram() + 
  labs(title = "Tuition")

cowplot::plot_grid(p3,p2,p1, ncol = 2)
```

---

### 1. EDA: Graduation Rate by Year   

- Let's look at the graduation rate over time for 20 randomly selected colleges

```{r echo = F, fig.height = 5}
set.seed(1234)
names <- college %>% distinct(instname) %>% sample_n(20) %>% pull()
college_samp <- college %>% filter(instname %in% names)

ggplot(data = college_samp, aes(x = year02, y = rate)) +
  geom_point() + 
  geom_line() +
  labs(x = "Years since 2002", 
       y = "Grad Rate", 
       title = "Graduation Rate over Time", 
       subtitle = "Faceted by College") +
  facet_wrap(~instname) +
  theme_bw()
```

---

### 1. EDA: Bivariate Analysis in 2002

```{r echo = F, fig.height = 5}
p1 <- college %>% 
  filter(year02 == 0) %>%
  ggplot(aes(x = faculty, y = rate)) +
  geom_point(alpha = 0.5) +
  labs(title = "Grad. Rate vs. Faculty in 2002")

p2 <- college %>% 
  filter(year02 == 0) %>%
  ggplot(aes(x = tuition, y = rate)) +
  geom_point(alpha = 0.5)  +
  labs(title = "Grad. Rate vs. Tuition in 2002")

cowplot::plot_grid(p1, p2, ncol = 2)
```

---

### 1. EDA: Bivariate Analysis in 2009

```{r echo = F, fig.height = 5}
p1 <- college %>% 
  filter(year02 == 7) %>%
  ggplot(aes(x = faculty, y = rate)) +
  geom_point(alpha = 0.5) +
  labs(title = "Grad. Rate vs. Faculty in 2009")

p2 <- college %>% 
  filter(year02 == 7) %>%
  ggplot(aes(x = tuition, y = rate)) +
  geom_point(alpha = 0.5)  +
  labs(title = "Grad. Rate vs. Tuition in 2009")

cowplot::plot_grid(p1, p2, ncol = 2)
```

---

### 2. Unconditional Means Model

- In an <font class="vocab">unconditional means model</font>, there are no predictors at any level 

- The goal of this model is to compare variability within colleges and variability between colleges 

.alert[
Let $Y_{ij}$ be the graduation rate of college $i$ in year $j$

$$Y_{ij} = \alpha_0 + u_i + \epsilon_{ij}\\[10pt]
u_i \sim N(0, \sigma^2_u) \hspace{10mm} \epsilon_{ij} \sim N(0, \sigma^2)$$
]

- $\sigma^2_u$: variability between colleges 

- $\sigma^2$: variability within a college 

---

### 2. Unconditional Means Model

We can fit the unconditional means model using <font class="vocab">`lmer`</font> function in the **lme4** package. 

```{r eval = F}
library(lme4)

model_0 <- lmer(rate ~ 1 + (1|instname), data = college)
summary(model_0)
```

---

### 2. Grad. Rates: Unconditional Means Model

```{r echo = F}
library(lme4)
model_0 <- lmer(rate ~ 1 + (1|instname), data = college)
summary(model_0)
```

---

- <font class="vocab">Coefficients</font>
    - $\hat{\alpha}_0 = 25.490$: mean graduation rate across all colleges
    
    - $\hat{\sigma}^2 = 9.938$: variance in within-school deviations between individual rate and college mean across all years

    - $\hat{\sigma}^2_u = 53.485$: variance in the between-college deviations between the college means and the overall mean across all colleges and all years

--

- <font class="vocab">Intraclass correlation</font>
$$\hat{\rho} = \frac{\hat{\sigma}^2_u}{\hat{\sigma}_u^2 + \hat{\sigma}^2} = \frac{53.485}{53.485 + 9.938} = 0.843$$

About `r round(53.485/(53.485 + 9.938) * 100, 1)`% of the total variation in graduation rates can be attributed to the difference among schools rather than the change over time within schools. We can also say the average correlation for any two responses from the same college is about `r round(53.485/(53.485 + 9.938), 3)`.

---

### 3. Unconditional growth model 

- In an <font class="vocab">unconditional growth model</font>,
time is added to the random effects (level one) model but no predictors in the fixed effects (level two) model

- The goal of this model is to determine how much of the within-school variability in graduation rate can be attributed to changes over time

- We can think of this as building individual models for the change in graduation rate over time for each college
    - We assume the same form of the relationship between `rate` and `year` for every college
    
.alert[
Let $Y_{ij}$ be the `rate` for college $i$ in year $j$

$$Y_{ij} = a_i + b_i \times \text{year02}_{ij} + \epsilon_{ij}\\
\epsilon_{ij} \sim N(0, \sigma^2)$$
]

---

### 3. Unconditional growth model 

.alert[
Let $Y_{ij}$ be the `rate` for college $i$ in year $j$

$$Y_{ij} = a_i + b_i \times \text{year02}_{ij} + \epsilon_{ij}\\
\epsilon_{ij} \sim N(0, \sigma^2)$$
]

- $a_i$: expected graduation rate for college $i$ at time 0 (2002)

- $b_i$: slope for college $i$, i.e. the rate of change in graduation rate for college $i$ over the time period

- $\epsilon_{ij}$: deviation in college $i$'s expected and actual graduation rate at time $j$
    - $\sigma^2$ is the variability in the $\epsilon_{ij}$'s
    
---

## Duke: Graduation rate over time.

```{r echo = F}
college %>%
  filter(instname == "Duke University") %>%
  ggplot(aes(x = year02, y = rate)) +
           geom_point() + 
            geom_line() + 
           labs(x = "Years Since 2002") +
  expand_limits(y = 0)
```

---

### 3. Unconditional growth model 

We will let $a_i$ and $b_i$ vary by college, so we can fit Level Two models that incorporate college-level variables to estimate these values 


.alert[
Let $Y_{ij}$ be the graduation rate for college $i$ in year $j$

**Level One**

$$Y_{ij} = a_i + b_i \times \text{year02}_{ij} + \epsilon_{ij}$$


**Level Two**
$$a_i = \alpha_0 + u_i \\
b_i = \beta_0 + v_i$$
]

---

.alert[
Let $Y_{ij}$ be the graduation rate for college $i$ in year $j$

**Level One**

$$Y_{ij} = a_i + b_i \times  \text{year02}_{ij} + \epsilon_{ij}$$


**Level Two**
$$a_i = \alpha_0 + u_i \\
b_i = \beta_0 + v_i$$
]

where $\epsilon_{ij} \sim N(0, \sigma^2)$ and 

$$\bigg[\begin{matrix}u \\ v  \end{matrix}\bigg] \sim N\Bigg(\bigg[\begin{matrix}0 \\ 0 \end{matrix}\bigg], \bigg[\begin{matrix}\sigma^2_u &  \rho\sigma_{u}\sigma_v \\ \rho\sigma_{u}\sigma_v & \sigma^2_{v} \end{matrix} \bigg] \Bigg)$$

---

### 3. Unconditional growth model 

- $\alpha_0$: mean graduation rate for all colleges in 2002

- $\beta_0$: mean yearly change in graduation rate for all colleges during the time period

--

- $\sigma^2$: within-school variability 

- $\sigma^2_u$: variability between colleges in the 2002 graduation rates

- $\sigma^2_v$: variability in the rate of change in the graduation rate (i.e. the slopes) 2002 - 2009

- $\sigma^2_u$ and $\sigma^2_v$ make up the between-school variability 

--

- $\rho_{uv}$: Correlation between a college's graduation rate in 2002 and the rate of change of the graduation rate 2002 - 2009


---

### Graduation rate: Unconditional growth model 

```{r echo = T, eval = F}
library(lme4)
model_1 <- lmer(rate ~ 1 + (year02|instname), 
                data = college)
summary(model_1)
```

---

```{r echo = F, eval = T}
library(lme4)
model_1 <- lmer(rate ~ year02 + (year02|instname), data = college)
summary(model_1)
```

---

### Understanding the model 

What do each of the following values tell you?

- $\hat{\alpha}_0 = 24.98$: 

- $\hat{\beta}_0 = 0.14$: 

- $\hat{\sigma}^2 = 7.74$: 

- $\hat{\sigma}^2_u = 59.10$: 

- $\hat{\sigma}^2_v= 0.48$: 

- $\rho_{uv} = -0.433$: 

---

### 4. Add predictors

- Do `faculty` and `tuition` affect graduation rates? 

- We will add these predictor variables to the fixed effects (level two) model, since they differ by college but don't change within a college in our data

.alert[
Let $Y_{ij}$ be the `rate` for college $i$ in year $j$

**Level One**

$$Y_{ij} = a_i + b_i \times \text{year02}_{ij} + \epsilon_{ij}$$


**Level Two**
$$a_i = \alpha_0 + \alpha_1 \times \text{faculty}_i + \alpha_2 \times \text{tuition}_i + u_i \\
b_i = \beta_0 + \beta_1 \times  \text{faculty}_i + \beta_2 \times \text{tuition}_i + v_i$$
]

---

### 4. Add predictors 

```{r echo = T, eval = F}
library(lme4)
model_2 <- lmer(rate ~ faculty + tuition  + year02 +
                  faculty:year02 + tuition:year02 +
                  (year02|instname), data = college)
summary(model_2)
```

---

```{r echo = F, eval = T}
model_2 <- lmer(rate ~ faculty + tuition  + year02 +
                  faculty:year02 + tuition:year02 +
                  (year02|instname), data = college)
summary(model_2)
```

---

## Write out the model 

---

### Predicted values 

We can use the `augment` function to get predicted values and residuals

```{r}
model_2_aug <- augment(model_2)
```

---

### Check Residuals

```{r echo = F, fig.height = 6}
p1 <- ggplot(data = model_2_aug, aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predicted", y = "Residuals")

p2 <- ggplot(data = model_2_aug, aes(x = faculty, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Faculty", y = "Residuals")

p3 <- ggplot(data = model_2_aug, aes(x = tuition, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Tuition", y = "Residuals")

p4 <- ggplot(data = model_2_aug, aes(x = year02, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Year since 2002", y = "Residuals")

cowplot::plot_grid(p1, p2, p3, p4, ncol = 2)
```

---

### Residuals 

```{r echo = F, fig.height = 6, warning = F, message = F}
p1 <- ggplot(data = model_2_aug, aes(x = .resid)) +
  geom_histogram() + 
  labs(title = "Distribution of Residuals")

p2 <- ggplot(data = model_2_aug, aes(sample = .resid)) +
  stat_qq() + 
  stat_qq_line() +
  labs(title = "Normal QQ Plot of Residuals")

cowplot::plot_grid(p1, p2, ncol = 2)
```

---

### Actual vs. Predicted Graduation Rates

```{r echo = F, fig.height = 6}
model_2_aug %>% filter(instname %in% c("Duke University", "University of North Carolina at Chapel Hill")) %>%
  ggplot() +
  geom_line(aes(x = year02, y = rate)) +
  geom_line(aes(x = year02, y = .fitted), color = "blue") +
  facet_wrap(~instname) +
  theme_bw() +
  expand_limits(y = 0) +
  labs(title = "Actual vs. Predicted Graduation Rates")
```

---

### References

*Broadening Your Statistical Horizons*

- ["Introduction to Multilevel Models"](https://bookdown.org/roback/bookdown-bysh/ch-multilevelintro.html) -
- ["Two Level Longitudinal Data"](https://bookdown.org/roback/bookdown-bysh/ch-lon.html)

---

class: middle, center

### Congrats on completing STA 210! `r emo::ji("smile")` 