---
title: "Lab 06: Logistic Regression"
date: "due Wed, Mar 27 at 11:59p"
output: 
  tufte::tufte_html:
    css: "./sta210-labs.css"
    tufte_variant: "envisioned"
    highlight: pygments
    fig_height: 10
link-citations: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE, 
                      message=FALSE, 
                      eval=FALSE)
```

Over the past ten years, recommendation systems have become increasingly popular as more companies strive to offer customized user experiences. Amazon recommends products you may like based on your browse and purchase history, Netflix recommends movies and TV shows based on your past viewing history, and music platforms like Spotify recommend songs you may like based on your listening history. While these recommendation systems can be built based on various different algorithms, they are all trying to achieve the same goal: use the characteristics of the products/movies/music a user is known to like to figure out the products/movies/music the user may like but hasn't discovered yet. 

In today's lab, we will focus on using song characteristics to determine whether or not a user will like a song. We will use logistic regression to build a model that predicts the probability a user likes a song using the relevant characteristics of the song. The data in this lab is from the [Spotify Song Attributes](https://www.kaggle.com/geomack/spotifyclassification) data set in Kaggle. This data set contains song characteristics of 2017 songs and whether or not the person liked the song. This data set contains the song preferences of a single person; therefore, our the scope of our analysis is limited to the person who created the data set. 


See ["How Does Spotify Know You So Well?"](https://medium.com/s/story/spotifys-discover-weekly-how-machine-learning-finds-your-new-music-19a41ab76efe) on Medium.com for more information about Spotify's recommendation algorithms.

# Getting Started 

- Go to the STA210-Sp19 organization on GitHub ([https://github.com/STA210-Sp19](https://github.com/STA210-Sp19)). Click on the repo with the prefix **lab-07-**. It contains the starter documents you need to complete the warmup exercise. 

- Clone the repo and create a new project in RStudio Cloud. 

- Configure git by typing the following in the **console**.

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="your name", user.email="your email")
```

```{marginfigure}
When configuring Git, be sure to use the email address that is associated with your GitHub account.
```

## Password caching

If you would like your git password cached for a week for this project, type the following in the **Terminal**:

```{bash eval=FALSE}
git config --global credential.helper 'cache --timeout 604800'
```

You will need to enter your GitHub username and password one more time after caching the password. After that you won't need to enter your credentials for 604800 seconds = 7 days.


## Packages

You will need the following packages for today's lab: 

```{r load-packages}
library(tidyverse)
library(broom)
library(arm)

#Fill in other packages as needed
```

## Project name: 

Currently your project is called *Untitled Project*. Update the name of your project to the title of today's lab.

# Warm up

Before we introduce the data, let's warm up with a simple exercise.

## YAML: 

- Pick one team member to update the *author* and *date* fields at the top of the R Markdown file. Knit, commit, and push all the updated documents to Github.

- Now, the remaining team members who have not been concurrently making these changes on their projects should click on the **Pull** button in their Git pane and observe that the changes are now reflected on their projects as well.

## Data

You will use data `spotify.csv` to build the logistic regression model and test the model the songs in `test_songs.csv`. 
Download `spotify.csv` [here](https://www2.stat.duke.edu/courses/Spring19/sta210.001/labs/data/spotify.csv) and `test_songs.csv` [here](https://www2.stat.duke.edu/courses/Spring19/sta210.001/labs/data/test_songs.csv) and upload the files to the data folder in your `lab-07` repo. Use the `read_csv` function to load the data. 

Data dictionary: 

**FILL IN**

# Exercises

## Exploratory Data Analysis


1. Use the `glimpse` function to see the size of the data set and the variable names. The response variable for this analysis is `like`, where 1 indicates that the user likes the song and 0 otherwise. The remaining variables will be considered as potential explanatory variables. 

The variable definitions can be found in the <a href="https://developer.spotify.com/documentation/web-api/reference/tracks/get-audio-features/" target="_blank">Spotify documentation page</a>. Read through the documentation to see the definitions of the variables used in this analysis. Based on the documentation, what is the definition of **`valence`**?


2. For now, we will use boxplots to look at the distributions of the numerical explanatory variables based on whether or not the Spotify user liked the song.

Before making the boxplots, change the response variable `like` so that it is a factor variable type.

Create boxplots of the variables `acousticness`, `danceability`, `duration_ms`,`energy`, and `instrumentalness`, each versus the response variable `like`. Use the `plot_grid()` function to fit the boxplots on one page by organizing them into three columns. An outline of the code is shown below: 
  
```{r,eval=F}
p1 <- ggplot(data=music,aes(x=like,y=acousticness)) + geom_boxplot() + 
  labs(title="Acousticness by Like")

# include code for plots p2 - p5

plot_grid(p1,p2,p3,p4,p5,ncol=3)
```

3. Create boxplots of the variables `liveness`, and `loudness`, `speechiness`, `tempo`, and `valence`, each versus the response variable `like`. Use the `plot_grid` function to fit the boxplots on one page by organizing them into two columns.

4. We will use bar graphs to display the distributions of the remaining variables, since they are best understand as describing categories rather than numerical data. Create bar graphs for the variables `key`, `mode`, and `time_signature`. As before, use `plot_grid` to neatly organize the plots into two columns. 


There are many ways to plot bar graphs. Below is an example of one way to plot a stacked bar graph.

```{r eval=F}
p1 <- ggplot(data=music,aes(x=key)) + geom_bar(aes(fill=like)) + 
  labs(title="Like vs.Key")
```

5. Examine the plots from the previous exercises. Based on these plots, which variables might be useful in differentiating which songs the user likes vs. those the user doesn't like? Briefly explain. 

## Part II: Logistic Regression Model

6. We will begin by considering the model that includes all the explanatory variables. Use the `glm()` function to build a logistic regression model that includes all the explanatory variables. An outline of the `glm` syntax is shown below. 

Display your model. Recall that the primary goal of this analysis is to predict the probability this Spotify user will like a song. Should we use this model for prediction? Briefly explain.


7. To enable better prediction, we would like to select only the most relevant variables for the model. We will use the `step` function in R, since it is able to accept `glm` objects. Conduct **backward selection** using the `step()` function with your model from Question 3 as the starting point. 

8. Display the model selected by backward selection.

9. There may be some variables that were selected by backward selection that do not significantly affect the probability of liking a song after accounting for the other variables in the model. Using a threshold of 0.05, remove any such variables (if applicable) and  display the final model. (*Note: Since we're using the p-values to determine if variables should be removed from the model, we are assuming the model is valid for inference. We will check the assumptions  to assess the model's adequancy in Part III.*)

10. We will now look at the variable `duration_ms`. 
    - What are the units of the variable `duration_ms`? </li>
    - Use the code below to obtain the 95% confidence interval for the coefficient of `duration_ms` in terms of its effect on the **<u>log-odds of liking a song</u>**.

```{r eval=F}
#confint.default ensures the confidence interval is calculated correctly for logistic model
confint.default(final.model,level=0.95)
```

11. Interpret the coefficient of `duration_ms` in terms of its effect on the **<u>odds of liking a song</u>**.

12. What would be the coefficient of `duration` if the units were changed to *seconds*?


## Part III: Model Assessment

13. We begin by examining the raw residuals vs. the predicted values. 
    - Use the `mutate()` function to add the predicted values and residuals to the `music` data set. You can use the code below to calculate the predicted values and residuals in terms of the predicted probability of liking a song.
    - Plot the residuals vs. the predicted values.
    - Describe the plot from part (b). Why is it unhelpful to use a plot of the raw residuals to assess the fit of a logistic model?

14. Instead of the raw residuals, we can use plots of the binned residuals to check the adequacy of the logistic model. See the lecture notes to recall the description of binned residual plots.
    - Use the code below to plot the binned residuals vs. the predicted values.

Similar to multiple linear regression, we need to plot the residuals vs. the explanatory variables. There are a lot of variables in our model, so we need to make the binned residuals plots in a way that will be easy for the reader to see. The plots produced by the `binnedplot` function are not the usual plot objects that can be used with the `plot_grid` function, so we must find another way to make organize multiple plots in a useful way.  

Put the code to create all the plots for the binned residuals vs. the explanatory variables in a single R code chunk. In order to put many plots on a single page, we will use the following header in the code chunk: 

15. Based on the plots from the previous exercise, is this logistic model a good fit? Briefly explain.

16. Look at the ROC curve and change in deviance statistic.

## Part IV: Prediction 


17. We finish this analysis by returning the original goal of determining whether or not the user would like a song. To test this, we will use our model to calculate the probability the user likes two songs: 

    - "Sign of the Times" by Harry Styles
    - "Hotline Bling" by Drake

The data for the songs can be found in *test_songs.csv*. Load the data into your R markdown file. 

The user likes "Hotline Bling" but doesn't like "Sign of the Times". Based on these two songs, how well did the model distinguish between songs this Spotify user likes vs. those he doesn't? Briefly explain. 


