---
title: "Lab 03: Conquering Merge Conflicts"
author: ""
date: "Fall 2019"
output: 
  html_document: 
    fig_height: 4
    fig_width: 9
---

```{r load-packages, include=FALSE}
library(tidyverse)
```

<font size="3">
1. **Everyone**: Clone your assignment repo in RStudio Cloud (`merge-conflicts-TEAMNAME`), 
and open the R Markdown file.

2. Assign the numbers 1, 2, 3, and 4 to each of the team members.
    - If you have 3 team members, make one person *Member 2* and *Member 4*.

3. Take turns in completing the steps below exactly as written, only <u>one member</u> at a time:

    - **Member 1**: Change the team name to your team name, knit, commit, push.
    
    - **Member 2**: Change the team name to something else, knit, commit, push. You 
will get an error. Pull. Review the document with the merge conflict. Resolve 
the conflict with the preferred change. Commit and push.

    - **Member 3**: Use the **diamonds** dataset to create a plot displaying the relationship between `price` and `carat`, 
controlling for `cut` of the diamonds dataset (see below for the plot). Knit, 
commit, push. You will get an error, read it, and pull. No merge conflicts should 
occur. Now push.

    - **Member 4**: Use the **diamonds**  dataset to create a plot displaying the relationship between `price` and `carat`, 
controlling for `cut` of the diamonds. Set `alpha = 0.5` in `geom_point()`. Knit, commit, push. You will get an error. Pull. Review the document with the merge conflict. Clear the merge conflict by choosing the correct/preferred change. Commit, and push.

4.  **All members:** Pull, and observe the changes in your document.

</font>
<br><br>

### Plot for Member 3 to create

```{r echo=FALSE}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point()
```