---
title: "Lab 02: Simple Linear Regression"
date: "due Wed, Jan 30 at 11:59p"
output: 
  tufte::tufte_html:
    css: "./sta210-labs.css"
    tufte_variant: "envisioned"
    highlight: pygments
    fig_height: 10
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE, 
                      message=FALSE)
```

The primary goal of today's lab is to give you practice with some of the tools you will need to conduct regression analysis using R. An additional goal for today is for you to be introduced to your teams and practice using GitHub and RStudio for collaboration. 

# Getting Started

Each of your assignments will begin with the following steps. You saw your TA demonstrate these steps, and they're outlined in detail here again. Going forward each lab will start with a "Getting started" section but details will be a bit more sparse than this. You can always refer back to this lab for a detailed list of the steps involved for getting started with an assignment.

## Clone Assignment Repo

- Go to the STA210-Sp19 organization on GitHub ([https://github.com/STA210-Sp19](https://github.com/STA210-Sp19)). Click on the repo with the prefix **lab-02-slr**. It contains the starter documents you need to complete the lab.

- Click on the green **Clone or download** button, select **Use HTTPS** (this might already be selected by default, and if it is, you'll see the text **Clone with HTTPS** as in the image below). Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the STA 210 course workspace. Create a **New Project from Git Repo**. You will need to click on the down arrow next to the **New Project** button to see this option.

- Copy and paste the URL of your assignment repo into the dialog box.  Be sure "Add packages from the base project" is checked. 

- Click OK, and you should see the contents from your GitHub repo in the **Files** pane in RStudio. 

## Configure git

There is one more piece of housekeeping we need to take care of before we get started. Specifically,  we need to configure your git so that RStudio can communicate with GitHub. This requires two pieces of information: your email address and your name.

To do so, you will use the `use_git_config()` function from the `usethis` package. 

Type the following lines of code in the **console** in RStudio filling in your name and email address.

```{marginfigure}
Your email address is the address tied to your GitHub account and your name should be first and last name.
```

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="your name", user.email="your email")
```

## Packages

We will use the following packages in today's lab. 

```{r load-packages}
library(tidyverse)
library(readr)
library(skimr)
library(broom)
library(knitr)
library(rcfss)
```

## Project name: 

Currently your project is called *Untitled Project*. Update the name of your project to be "Lab 02 - Simple Linear Regression"

# Warm up

**Pick one team member to complete the steps in this section while the others contribute to the discussion but do not actually touch the files on their computer.**

Before we introduce the data, let's warm up with some simple exercises.

## YAML: 

Open the R Markdown (Rmd) file in your project, change the author name to your **team** name, and knit the document.

## Commiting and pushing changes:

- Go to the **Git** pane in your RStudio. 
- View the **Diff** and confirm that you are happy with the changes.
- Add a commit message like "Update team name" in the **Commit message** box and hit **Commit**.
- Click on **Push**. This will prompt a dialogue box where you first need to enter your user name, and then your password.

## Pulling changes:

Now, the remaining team members who have not been concurrently making these changes on their projects should click on the **Pull** button in their Git pane and observe that the changes are now reflected on their projects as well.

## Data 

In today's lab, we will analyze the `scorecard` dataset from the rcfss package. This dataset contains information about a sample of colleges from the Department of Education's College Scorecard. Type `?scorecard` in the **console** to learn more about the dataset and the definition of each variable. Today's analysis will focus on the following variables: 

|  |  |
|-----------|------------------------------------------|
| `type` | Type of college (Public, Private - nonprofit, Private - for profit) |
| `cost` | The average annual cost of attendance, including tuition and feeds, books and supplies, and living expenses, minus the average grant/scholarship aid |
| `admrate` | Undergraduate admissions rate (from 0 - 100%) |

# Exercises 

1. Plot a histogram to examine the distribution of `admrate`. What is the shape of the distribution? 

2. To better understand the distribution of `admrate`, we would like calculate measures of center and spread of the distribution. Fill in the code below to use the `skim` function to calculate summary statistics for the variable. Report the appropriate measures of center (mean or median) and spread (standard deviation or IQR) based on the shape of the distribution from Question 1.

```{r eval=FALSE}
scorecard %>%
  select(admrate) %>%
  skim
```
3. Plot the distribution of `cost` using a different plot type (i.e. not a histogram) from the [ggplot2 Cheat Sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf). In addition to the scatterplot, use the `skim` function to calculate summary statistics for `cost`. 

Use the histogram and summary statistics to describe the distribution of `cost`. 

4. One nice feature of the `skim` function is that it provides information about the number of observations that are missing values of the variable. How many observations have missing values of `admrate`? How many observations have missing values of `cost`? 

5. Later in the semester, we will techniques to deal with missing values in the data. For now, however, we will only include complete observations in the remainder of the analysis. We can use the `filter` function to select only the rows that values for both `cost` and `admrate`. 

Fill in the code below to create a new dataset called `scorecard_new` that only includes complete observations, i.e. those with values for both `admrate` and `cost`.

```{r eval=FALSE}
__________ <- scorecard %>%
  filter(!is.na(admrate),________)
```

```{marginfigure}
Learn more about the `filter` function in [Section 5.2 of R for Data Science] (https://r4ds.had.co.nz/transform.html#filter-rows-with-filter)
```

6. Create a scatterplot to display the relationship between `cost` (response variable) and `admrate` (explanatory variable). Use the scatterplot to describe the relationship between the two variables. The scatterplot should have labeled axes as well as an informative title. 

7. The data contains information about the type of college, and we would like to see the relationship while also incorporating information about the type of college. One way to do this is to use a different color for each type of college. Fill in the code below the scatterplot from Question 3 with the marker colors based on the variable `type`. Briefly describe how the relationship between `admrate` and `cost` compare across values of `type`. 

```{r eval=FALSE}
ggplot(data=scorecard, mapping=aes(x=admrate, y=cost, color=type)) + 
  _____________________
```


8. Fit a regression model to describe the relationship between a college's admission rate and cost. Use the `tidy` function to display the estimated coefficients and associated test statistic, p-value and 95% confidence interval. 

9. Interpret the slope in the context of the problem. Does the intercept have a meaningful interpretation? If so, write the interpretation in the context of the problem. Otherwise, explain why the interpretation is not meaningful. 

10. While the `tidy` function is used to create a data frame of the coefficients and their associated inferential statistics, we can get a one-row summary of the model using the `glance` function. Use the `glance` function to get a summary of the model fit in Exercise 8. 

```{marginfigure}
See the [documentation for `glance`](https://rdrr.io/cran/broom/man/glance.lm.html) to see the model summary values produced by `glance`. 
```

11. What is the value of $R^2$? Interpret this value in the context of the problem. 

12. What is the value of $\hat{\sigma}^2$, the estimated regression variance? 

13. What is the 95% confidence interval for the coefficient of `admrate`, i.e. the slope? Interpret the interval in the context of the data. 

14. To conduct inference for a population slope, we will test the following hypotheses: 

$$H_0: \beta_1 = 0 \hspace{5mm} \text{versus} \hspace{5mm} \beta_1 \neq 0$$

State what the null and alternative hypotheses mean in terms of the relationship between `admrate` and `cost`.

15. Consider the confidence interval from Question 13 and the hypotheses in Question 14. Is the confidence interval consistent with the null or alternative hypothesis? Briefly explain.  


*You're done! Commit all remaining changes, use the commit message "Done with Lab 2!", and push. Before you wrap up the assignment, make sure the .Rmd, .html, and .md documents are all updated on your GitHub repo.*
