---
title: "Lab 03: Dealing with Merge Conflicts"
author: ""
date: "Fall 2019"
output: 
  html_document: 
    fig_height: 4
    fig_width: 9
---

<style>
ul{
  padding: 10pt;
}

ol{
padding: 15pt;
}
</style>

```{r load-packages, include=FALSE}
library(tidyverse)
```

<font color = "blue">**Step 1**</font>: **Everyone**: Clone your assignment repo in RStudio Cloud (`merge-conflicts-TEAMNAME`), 
and open the R Markdown file. Don't forget to configure git! 

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="GITHUB USERNAME", user.email="your email")
```
<br>

<font color = "blue">**Step 2**</font>: Assign the numbers 1, 2, 3, and 4 to each of the team members. **<i>Member 4</i>** should look at the group's repo on GitHub.com to ensure that the other members' files are pushed to GitHub after every step. 


<font color = "blue">**Step 3**</font>: Take turns in completing the steps below exactly as written, only <u>**one member**</u> at a time:

- **<i>Member 1</i>**: Change the team name to your team name. Knit, commit, and push.

- **<i>Member 2</i>**: Change the team name something different, i.e. not your team name. Knit, commit, push. You will get an error. 

  - Pull and review the document with the merge conflict. Show the error to your teammates.
  - A merge conflict occurred because you edited the same part of the document as Member 1. Resolve the conflict with whichever name you want to keep. 
  - Knit, commit, and push.

- **<i>Member 3</i>**: Use the **diamonds** dataset to create a plot displaying the relationship between `price` and `carat`, controlling for `cut` of the diamonds dataset (see below for the plot). *Note: You can control fro the cut by using the "color" option in the `aes()`.*

  - Knit, commit, push. You will get an error. 
  - Read it and pull. Show the error your teammates.
  - No merge conflicts should occur, since you edited a different part of the document from Members 1 and 2.
  - Knit, commit, and push. 


<font color = "blue">**Step 4**</font>: **<i>All members</i>:** Pull, and observe the changes in your document.
<br><br>

### Plot for Member 3 to create

```{r echo=FALSE}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point()
```