---
title: "Lab 03: Dealing with Merge Conflicts"
author: ""
date: "Fall 2019"
output: 
  html_document: 
    fig_height: 4
    fig_width: 9
---

```{r load-packages, include=FALSE}
library(tidyverse)
```

<font size="3">
1. **Everyone**: Clone your assignment repo in RStudio Cloud (`merge-conflicts-TEAMNAME`), 
and open the R Markdown file. Don't forget to configure git! 

```{r eval = F}
configureation
```

2. Assign the numbers 1, 2, 3, and 4 to each of the team members.
    - If you have 3 team members, make one person *Member 2* and *Member 4*.

3. Take turns in completing the steps below exactly as written, only <u>one member</u> at a time:

    - **Member 1**: Change the team name to your team name, knit, commit, push.
    
    - **Member 2**: Change the team name something different than your team name. Knit, commit, push. You 
will get an error. 
    - Pull.
    - Review the document with the merge conflict. 
    - Resolve the conflict with whichever name you want to keep. 
    - Knit, commit and push.

    - **Member 3**: Use the **diamonds** dataset to create a plot displaying the relationship between `price` and `carat`, controlling for `cut` of the diamonds dataset (see below for the plot). *Note: You can control fro the cut by using including "color" in the `aes()`.*
    - Knit, commit, push. 
    - You will get an error, read it, and pull. 
    - No merge conflicts should occur, since you edited a different part of the document from Members 1 and 2.
    - Now push.

4.  **All members:** Pull, and observe the changes in your document.

</font>
<br><br>

### Plot for Member 3 to create

```{r echo=FALSE}
ggplot(data = diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point()
```