---
title: "Lab 03: Inference for Simple Linear Regression"
subtitle: "Total: 75 points"
date: "due Tuesday, September 17 at 11:59p"
output: 
  tufte::tufte_html:
    css: "./sta210-labs.css"
    tufte_variant: "envisioned"
    highlight: pygments
    fig_height: 10
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE, 
                      message=FALSE, 
                      eval = FALSE)
```

The primary goal of todays' lab is to practice statistical inference for simple linear regression. Additionally, we will explore more data visualization and wrangling options as we conduct the exploratory data analysis. 

# Getting Started

Each of your assignments will begin with the following steps.

## Clone Assignment Repo

- Go to the sta210-fa19 organization on GitHub ([https://github.com/sta210-fa19](https://github.com/sta210-fa19)). Click on the repo with the prefix **lab-03-**. It contains the starter documents you need to complete the lab.

- Click on the green **Clone or download** button, select **Use HTTPS** (this might already be selected by default, and if it is, you'll see the text **Clone with HTTPS** as in the image below). Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the STA 210 course workspace. Create a **New Project from Git Repo**. You will need to click on the down arrow next to the **New Project** button to see this option.

- Copy and paste the URL of your assignment repo into the dialog box.  Be sure "Add packages from the base project" is checked. 

- Click OK, and you should see the contents from your GitHub repo in the **Files** pane in RStudio. 

## Configure git

There is one more piece of housekeeping we need to take care of before we get started. Specifically,  we need to configure your git so that RStudio can communicate with GitHub. This requires two pieces of information: your email address and your name.

To do so, you will use the `use_git_config()` function from the `usethis` package. 

Type the following lines of code in the **console** in RStudio filling in your name and email address.

```{marginfigure}
Your email address is the address tied to your GitHub account and your name should be your GitHub username.
```

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="GITHUB USERNAME", user.email="your email")
```

## Packages

We will use the following packages in today's lab. 

```{r load-packages}
library(tidyverse)
library(knitr)
library(broom)
library(openintro) 

#fill in any other packages you use in the lab
```

## Project name: 

Currently your project is called *Untitled Project*. Update the name of your project to be "Lab 03 - Inference for SLR".

# Warm up

Complete the [Lab 03: Dealing with Merge Conflicts](https://www2.stat.duke.edu/courses/Fall19/sta210.001/labs/lab-03-merge-conflicts.html) exercise before starting this lab. 

# Data 

Though Starbucks is most famous for it's ever-growing selection of coffee drinks, it has quite the [selection of food](https://www.starbucks.com/promo/food) as well. In today's lab, we will analyze the nutritional data of 77 food items sold at Starbucks. The data was originally obtained from the Starbucks menu in 2011; however, many of the items are still available today (and haven't had much change in their nutritional values!). 

The data is available in the `starbucks` dataset from the openintro package. It contains the following variables:  


| Variable 	| Description 	|
|------------	|----------------------------------------------------	|
| `item` 	| Name of food item 	|
| `calories` 	| Total number of calories 	|
| `fat` 	| Total fat (in grams) 	|
| `carb` 	| Total carbohydrate (in grams) 	|
| `fiber` 	| Total fiber (in grams) 	|
| `protein` 	| Total protein (in grams) 	|
| `type` 	| Food category (e.g. bakery, sandwich, salad, etc.) 	|

# Exercises 

When you walk to the counter at Starbucks, you'll notice the large display of its most popular food items. Often, the number of calories is shown in the display case but no other nutritional information is visible. Therefore, we'd like to fit a regression model so we can use the number of calories in a food item to estimate its other nutritional values. In this lab, we'll focus on the total carbohydrates (`carb`).

## Exploratory Data Analysis

1. What is the predictor variable? What is the response variable? 

2. Let's begin by examining the distribution of the **<i>response variable</i>**. Make a histogram to display the distribution of the response variable. Describe the shape of the distribution. 

*Note: Inlcude an informative title and informative labels for the x and y axes. This applies to all plots in the lab.*

3. Previously, we have used the `skim` function to quickly generate summary statistics for the variables in the dataset. We can also use the `summarise` function to calculate summary statistics from the data. By using the `summarise` function, you can choose which summary statistics you'd like to include.

Use the `summarise` function to calculate measures of center and spread for the **<i>response variable</i>**. Only include the measures of center and spread that are appropriate for describing the distribution of the response.

```{marginfigure}
See the [dplyr reference page](https://dplyr.tidyverse.org/reference/summarise.html) for more information about the `summarise` function. 
```

Below is example code for finding the minimum value of the response.

```{r}
starbucks %>%
  summarise(min_val = min(response))
```

4. Next, we want to examine the distribution of the predictor variable. We've primarily used histograms to visualize quantitative data, but let's try something new! Plot the distribution of the **<i>predictor variable</i>** using an appropriate plot that is **not** a histogram. You can use the [ggplot reference page](https://ggplot2.tidyverse.org/reference/index.html) to help generate ideas and see example code.

5. Briefly describe the type of plot you chose and what features of the distribution are visible using that plot. For example, the description of the histogram may be, *A histogram divides the range of the variable into bins with equal width and displays the number of observations in each bin. From a histogram, one can clearly see the shape of the data. One can also get an idea of the approximate center and spread of the data.*

6. Compare the features of the distribution that are visible from your plot versus a histogram. Which type of plot do you think is more effective for visualizing the distribution of a quantitative variable? Briefly explain your choice.

7. Make a plot displaying the relationship between the response and predictor variables. Describe the relationship between the two variables.  

8. From the plot in the previous question, what assumption is potentially violated? Briefly explain your reasoning. *Note: We still need to examine the residuals before making a final determination about the model assumptions; however, we can start to get intuition using during the exploratory data analysis. 

## Regression

9. Fit the regression model and display the output including the 95% confidence interval for the slope. Write the model equation. *Use words/variable names when you write the equation (not "x" and "y").* 

10. Recall the assumption you mentioned in Exercise 8. Make the appropriate residual plot(s) to assess that assumption. What do you conclude about whether this model assumption is satisfied? Briefly explain your reasoning. 

*Note: To keep the lab manageable, we will not check the remaining assumptions. In practice, however, you should check <u>all</u> assumptions for before using the model for statistical inference.*

11. Regardless of your answer to Exercise 10, we will proceed as if all assumptions are met. The estimated regression variance, $\hat{\sigma}^2$ is used to calculate the the standard error for the slope and intervals for predictions (recall the formulas for standard error in your notes). Fill in the name of your model to find $\hat{\sigma}^2$ using the `glance` function.  

```{r}
glance(_____)$sigma
```

Briefly describe what is being measured by  $\hat{\sigma}^2$ measures. *Hint: It may help to recall the [visual representation of simple linear regression](https://www2.stat.duke.edu/courses/Fall19/sta210.001/slides/lec-slides/03-slr.html#45). What takes the value $\sigma^2$ in the diagram?

## Statistical Inference
12. What is the 95% confidence interval for the slope? Interpret this interval in the context of the data.

13. Suppose we want to test the following hypotheses: 

$$\begin{aligned}&H_0: \beta_1 = 0 \\
&H_a: \beta_1 \neq 0
\end{aligned}$$

State the null and alternative hypotheses using words in the context of the data.

14. What is the p-value of the hypothesis test from the previous exercise? Use the p-value to state your conclusion from the test in the context of the data.

15. Consider the confidence interval from Exercise 12 and the hypotheses in Exercise 13. Is the confidence interval consistent with the null or alternative hypothesis? Briefly explain.

## Prediction 

16. You'd like to purchase a piece of pumpkin bread from Starbucks to celebrate fall! According to [Starbuck's website](https://www.starbucks.com/menu/food/bakery/pumpkin-bread-lb?foodZone=9999), pumpkin bread has 410 calories. Calculate the predicted mean carbohydrates for all pumpkin bread sold by Starbucks. Include the estimate and appropriate interval. 

*You're done! Commit all remaining changes, use the commit message "Done with Lab 3!", and push. Before you wrap up the assignment, make sure the .Rmd and .md documents are all updated on your GitHub repo. There is a 10% penality if the .md file is not updated on GitHub.*