---
title: "Extra Credit Assignment"
subtitle: "Matrix Approach to Linear Regression"
author: "due Mon, Mar 18 at 11:59p"
date: ""
output: 
  html_document:
    css: "./sta210-hw.css"
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE,
                      message=FALSE)
```

This extra credit assignment is to give you practice using the matrix equations to fit a regression model and explore some model diagnostics. See the [multiple regression lecture](https://www2.stat.duke.edu/courses/Spring19/sta210.001/slides/lec-slides/07-mlr-special-predictors.html#7) notes and the [supplemental notes on GitHub](https://github.com/STA210-Sp19/supplemental-notes) to review matrix equations and sample code.

**You must attempt every question to receive any extra credit.**

# Getting Started

- Go to the STA210-Sp19 organization on GitHub ([https://github.com/STA210-Sp19](https://github.com/STA210-Sp19)). 
- Create a new repo called **extra-credit-USERNAME** where, USERNAME is your GitHub Username. Make sure your repo is **private**.
- Clone the repo in RStudio Cloud. 
- In RStudio Cloud, create a new R Markdown file (*File -> New File -> R Markdown*). Call the RMarkdown file **extra-credit.Rmd**. Be sure to fill in an appropriate title, name, and date in the YAML header. 
- In the YAML header, change the output to **github_document**.
- Configure git using the `use_git_config()` function. You can also cache your password.


## Packages 

You will need the following 

```{r}
library(tidyverse)
library(Sleuth3)
```

## Data 

The data for this assignment is from the `case1101` data frame in the **Sleuth3** package. This data was originally collected to investigate why women seemingly have a lower tolerance for alcohol and develop more alcohol-related diseases than men. In this analysis; however, you will focus on the relationship between the first-pass metabolism of alcohol (a measure of the concentration of alcohol) and the activity of the alcohol-degrading enzymes in the stomach for men. The data frame contains the following variables: 

- `Metabol`: first-pass metabolism of alcohol in the stomach (in mmol/liter-hour)
- `Gastric`: gastric alcohol dehydrogenase activity in the stomach (in $\mu$mol/min/g of tissue)
- `Sex`: sex of the subject
- `Alcohol`: whether the subject is alcoholic

You can learn more about the data frame by viewing the help file (`??Sleuth3::case1101`).

# Questions

The following questions will walk you through a matrix approach to regression analysis. See the [multiple regression lecture](https://www2.stat.duke.edu/courses/Spring19/sta210.001/slides/lec-slides/07-mlr-special-predictors.html#7) notes and the [supplemental notes on GitHub](https://github.com/STA210-Sp19/supplemental-notes) to review matrix equations and sample code. You will use R to do all calculations. 

Just like a standard assignment, but sure to show your R code and output and include the interpretation/narrative when required. 

**To receive any extra credit, you must attempt every question and <u>all work</u> must be done using the matrix equations. You will not receive credit for using built-in R functions like `lm` and `augment` to fit the model and obtain diagnostics.**

You will only analyze data for males in this analysis. The response variable is `Metabol` and the predictor variables are `Gastric` and `Alcohol`. To help you get started, the code to filter the data and create $\mathbf{Y}$, the vector of responses, is below.

```{r}
alcohol_met <-  case1101 %>%
  filter(Sex == "Male")
```

```{r}
y <- alcohol_met %>%
  select(Metabol) %>%
  as.matrix()
```

1. Create the $\mathbf{X}$ matrix. *Hint: The predictor variable `Alcohol` is categorical, so you will need to create an indicator variable. Make an indicator variable for the level "Alcoholic".*

2. The columns of the $\mathbf{X}$ matrix correspond to the coefficients in the regression model. What model coefficient is represented in the first column of $\mathbf{X}$? 

3. Use the matrix equations to estimate the coefficients, $\hat{\boldsymbol{\beta}}$.
    - Interpret the coefficient for each predictor variable.

4. Calculate the predicted values, $\hat{\mathbf{Y}}$, and the residuals $\mathbf{e}$.  

5. Use the answer from the previous question to estimate $\sigma^2$, the regression variance. 

6. Calculate $\mathbf{H}$, the hat matrix. What do the diagonal elements of $\mathbf{H}$  represent?

7. Based on the hat matrix, are there any observations that stand out, i.e. observations of interest that you should further investigate? If so, which observation(s)? 
    - Include the criteria you used to make the determination.

8. Calculate the variance-covariance matrix for the coefficients. Are there concerns with multicollinearity? Briefly explain and show the calculations used to make the conclusion.

*Once you're finished, be sure to push your work to GitHub! Go to your GitHub repo to make sure all of your work submitted. The only work that will be considered for extra credit is the work in the .Rmd <u>and</u> .md file in the **extra-credit-** repo by the deadline.*


## Resources

Original data from M. Frezza et al., “High Blood Alcohol Levels in Women,” *New England Journal of Medicine*, 322 (1990): 95–99.


