---
title: "HW 01"
subtitle: "due Wed, Jan 29 at 11:59p"
author: ""
date: ""
output: 
  html_document:
    css: "./sta210-hw.css"
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE,
                      message=FALSE)
```
<br><br>

# Getting Started

A brief outline of getting started is shown below. See the [Lab 01 Instructions](https://www2.stat.duke.edu/courses/Fall19/sta199.001/labs/lab-01-intro-r.html) for more details about the steps. 

- Go to the sta210-fa19 organization on GitHub ([https://github.com/sta210-fa19](https://github.com/sta210-fa19)). Click on the repo with the prefix **hw-01**. It contains the starter documents you need to complete the lab.
- Clone the repo in RStudio Cloud
- Configure git using the `use_git_config()` function.

## Tips

Here are some tips as you complete HW 01:

- Periodically Knit your document and commit changes (the more often the better, for example, once per each new task)
- Push all your changes back to your GitHub repo. The Git pane in RStudio should be empty after you push. You can also check your assignment repo on github.com to make sure it has updated.
- Once you have completed your work, you will submit your assignment in Gradescope. You are welcome to resubmit multiple times until the assignment deadline. We will grade the most recent version of the .pdf file in Gradescope.

## Packages 

We will use the following packages in this assignment: 

```{r}
library(tidyverse)
library(modelr)
library(readr) 
library(broom)
library(knitr)
library(openintro)
```

If you need to install any of the packages, type `install.packages("package_name")` in the console, where `package_name` is the package you need to install. 

# Data - UPDATE THE VARIABLE DEFINTIONS!

For this assignment, we will be using the `smoking` dataset in the openintro R package. The data was originally provided by https://www.stem.org.uk/. 

This dataset contains smoking habits and other demogrpahic information for 1,691 randomly selected survey respondents in the United Kingdom. The following variables are in the dataset: 

|  |  |
|-----------|------------------------------------------|
| `family_income` | Family income of the student |
| `gift_aid` | Gift aid, in ($ thousands) |
| `price_paid` | Price paid by the student (= tuition - gift_aid) |



# Questions

## Part 1: Conceptual Problems

The *Conceptual Problems* section of homework contains short answer questions about the concepts discussed in class. Some of these questions may also require short chunks of code to produce the output needed to answer the question. Answers should be written in complete sentences.

1. According to a [2018 article by the BBC](https://www.bbc.com/news/uk-43316697), the average age of a resident in the UK is 40. *Is the average age of <u>smokers</u> in the UK significantly different from 40, the average age of all UK residents, 40?*?

    a. State the hypotheses both in words and using statistical notiation.  You can use the syntax \$mu\$ to produce $\mu$ when typing the hypotheses. ^[See [Getting Started with LaTex](https://artofproblemsolving.com/wiki/index.php/LaTeX:LaTeX_on_AoPS#Getting_Started_with_LaTeX) and [LaTex Symbols Guide](https://artofproblemsolving.com/wiki/index.php/LaTeX:Symbols#Greek_Letters) for more information about typing statistical notation using LaTex.]
    b. Are the conditions met to conduct a *t*-test? Include discussion about each condition, including an explanation about your conclusion and any supporting graphs, code, or output, as needed.
    c.  Use equations to calculate the test statistic. Show the code used to calculate the test statistic and the resulting output. Write the definition of the test statistic in the context of this problem. 
    d. What are the degrees of freedom associated with this test statistic?
    e. Calculate the p-value and use it to state your conclusion in the context of this problem. 
    f. If you were to construct a 95% confidence interval that corresponded to this hypothesis test, would you 40 to be in the interval? Briefly explain why or why not.


2. *Is there a significant difference in the average age of of people in the UK who smoke versus those who don't?* To answer this question, will use a *t*-test to compare the average age of those who smoke versus those don't. 

    a. State the hypotheses both in words and using statistical notation.
    b. Are the conditions met to conduct a *t*-test? Include discussion about each condition, including an explanation about your conclusion and any supporting graphs, code, or output, as needed.
    c. Use the `t.test` function to test the hypotheses from part(a). Write the definition of the test statistic in the context of this problem. 
    d. Write the definition of the p-value in the context of this problem.
    e. Use the p-value to make a conclusion in the context of this problem. 


## Part 2: Data Analysis

The *Data Analysis* section of homework contains open-ended data analysis questions. Your response should be neatly organized and should read as a complete narrative. This means that in addition to addressing the question(s) stated below, you should include exploratory data analysis and check the appropriate model assumptions. In short, these questions should be treated as "mini-projects".

9. Fit a regression model that could be used to predict the number of bike rentals based on the temperature during the **summer season**. Use your model to answer the following: 
    - Describe the relationship between the two variables including discussion about whether the relationship is statistically significant. 
    - In Washington D.C., the average temperature in the summer is about 30 degrees Celsius (about 87 degrees Fahrenheit). Use your model to predict (1) the average number of bike rentals for the days in the summer that are 30 degrees Celsius and (2) the number of bike rentals we can expect if it's 30 degrees on a single day in the summer.
    
*Notes:* 
*- Be sure to include all relevant code and resulting output.*
*- All plots should have proper labels for the axes and an informative title.* 

# Submitting Your Assignment



# Grading

| **Total** | **60** |
|-------------------------------------------------------------|:--:|
| Questions 1 - 8 | 30 |
| Question 9 | 20 |
| Documents neatly organized (.Rmd and .md files) | 5 |
| Narrative in full sentences and neatly organized| 3 |
| Regular and informative commit messages | 2 |
