---
title: "rstudio::conf"
subtitle: "due Fri, Jan 31 at 11:59p"
author: ""
date: ""
output: 
  html_document:
    css: "./sta210-hw.css"
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE,
                      message=FALSE)
```

```{r echo = F}
library(emo)
```

## Part 1: Watch a session from rstudio::conf

There are three ways to watch a session: 

### Option 1: Attend an on-campus rstudio::conf watch party

Find the full schedule and register for a session here: https://blogs.library.duke.edu/data/2020/01/13/2020-rstudio-conference-livestream-coming-to-duke-libraries/?aiEnableCheckShortcode=true

Space is limited, so register in advance! There will be swag from RStudio and some sessions include food. 


### Option 2: Watch rstudio::conf live online

You can live stream the conference using the following link: **ADD LINK**

### Option 3: Watch talks from 2019 rstudio::conf 

If you are unable to attend a watch party or watch a talk from the 2020 conference, you can watch one or two talks from the 2019 conference. 

- Find the talks from the 2019 event here: https://resources.rstudio.com/rstudio-conf-2019

- Watch either: 
    - 2 short talks (40 minutes or less) or
    - 1 long talk (over 30 minutes)
    
## Part 2: Submit a reflection 

Once you've watched rstudio::conf, submit a short reflection here: **ADD LINK**

For the email address, use NetID@duke.edu, so I can enter your reflection grade in Sakai. 


