---
title: "Calories vs. Alcohol in Beer"
author:  ""
date: "`r Sys.Date()`"
output: html_document
---

```{r global-options}
knitr::opts_chunk$set(warning=FALSE,
                   message=FALSE)
```

```{r}
library(tidyverse)
library(readr)
library(broom)
```

```{r load-data}
beer <- read_csv("data/beer.csv") %>%
  filter(!is.na(Carbohydrates), !is.na(PercentAlcohol))
```

In this analysis, we will analyze the relationship between the amount of alcohol (`PercentAlcohol`) and the caloric content (`CaloriesPer12Oz`) in domestic beers. Let `PercentAlcohol` be the predictor variable and `CaloriesPer12Oz` the response variable. 

Due to limited class time, we will not do the exploratory data analysis in this example. In practice, however, you should always start with the exploratory data analysis.

*You can add your answers to this R Markdown document*.

1. Calculate a regression model to describe the relationship between `PercentAlcohol` and `CaloriesPer12Oz`. Display the model output. 

```{r model}

```

2. Does it make sense to interpret the intercept? Why or why not?


3. Interpret the 95% confidence interval for the slope in the context of the data. 


4. Find the critical value, $t^*$, used to calculate the 95% confidence interval. The code below is a guide; uncomment and complete the lines of code to calculate and display the critical value.

```{r}
n <- nrow(beer)

#df <- _________  

#crit_val <- qt(_____,df)
#crit_val
```

5. Interpret the test statistic in the context of the data. 

6. How was the p-value calculated? Fill in the code below to calculate the p-value. The code below is a guide; uncomment and complete the lines of code to calculate and display the pvalue.

```{r}
#test_statistic <- _______

#prob <- 1 - pt(abs(test_statistic),df) #prob is the proportion of area to the right of the test statistic.


```

7. Fill in the code below to calculate the predicted calories and corresponding 90% interval for a single beer with alcohol content of 4.3%. 

```{r}
x0 <-  data.frame(PercentAlcohol=4.3)
#predict.lm(model,x0,interval="________",conf.level=________) #<<
```


8. Fill in the code below to calculate the predicted calories and corresponding 90% interval for the subset of beers with alcohol content of 4.3%. 

```{r}
x0 <-  data.frame(PercentAlcohol=4.3)
#predict.lm(model,x0,interval="________",conf.level=________) #<<
```