---
title: "Heart Weight vs. Body Weight in Cats"
author:  ""
date: "`r Sys.Date()`"
output: html_document
---

```{r global-options, echo = F}
knitr::opts_chunk$set(warning=FALSE,
                   message=FALSE, 
                   eval = FALSE)
```

```{r load-packages}
library(tidyverse)
library(broom)
library(MASS) #cats dataset
```

## Introduction 

In this analysis, we will analyze the relationship between the body weight (`Bwt`) and heart weight (`Hwt`) for 144 domestic cats. When veterinarians prescribe heart medicine for cats, they often need to calibrate the medicine to the weight of the heart. Obtaining a cat's heart weight is quite difficult, so veterinarians need to estimate it using more easily obtainable measurement, such as body weight. 

Therefore, the goal of today's analysis is to fit a regression model to describe the relationship between the body weight and heart weight in domestic cats. 

We will use the `cats` dataset from the MASS package. The variables in this dataset are 
- `Sex`: Male (M) or Female (F)
- `Bwt`: Body weight in kilograms (kg)
- `Hwt`: Heart weight in grams (g)


## Part 1: Exploratory Data Analysis

Before fitting the regression model, we need to do exploratory data analysis (EDA) to better understand the data. Below are some of the plots and summary statistics we can use for the EDA. When conducting a regression analysis, we need to include plots and narrative about what we learn in the EDA. Due to limited class time, we won't include the narrative here but will just display the plots. 

### Univariate Plots 
```{r univariate-plots}
p1 <- ggplot(data = cats, aes(x = Bwt)) + 
  geom_histogram(binwidth = 0.1) +
  labs(x = "Body Weight (kg)", 
       title = "Distribution of Body Weight")

p2 <- ggplot(data = cats, aes(x = Hwt)) + 
  geom_histogram(binwidth = 0.5) +
  labs(x = "Heart Weight (g)", 
       title = "Distribution of Heart Weight")

cowplot::plot_grid(p1, p2, ncol = 2) 
```

- There may be differences in the relationship of heart and body weight for male and female cats, so we'd like to know the distribution of `Sex` so we have full context when interpreting the results. 

```{r bar-plot}
ggplot(data = cats, aes(x = Sex)) +
  geom_bar() +
  geom_text(stat = "count", aes(label =..count..), vjust = 3, color = "white")
```

### Summary Statistics 

```{r summary-stats}
cats %>%
  skimr::skim()
```

### Scatterplot

```{r scatterplot}
ggplot(data = cats, aes(x = Bwt, y = Hwt)) +
  geom_point() + 
  labs(x = "Body Weight (kg)",
       y = "Heart Weight (g)", 
       title = "Heart Weight vs. Body Weight")
```


## Part 2: Fit the Model & Check Assumptions 

1. Fill in the code below to fit the regression model describing the relationship between a cat's body weight and heart weight. Display the output. 

```{r model}
bwt_hwt_model <- lm(___ ~ ____, data = cats) 

tidy(bwt_hwt_model, conf.int = TRUE, level = 0.95) %>%
  kable(digits = 3)
```


2. Write the model equation. 

3. Before doing any statistical inference or interpretation, let's check the model assumptions. Fill in the code below to calculate the residuals.  

```{r resid}
cats <- cats %>%
  mutate(resid = residuals(_____))
```

4. Use the code below to plot the residuals versus the predictor variable. Include an informative title and informative labels for the x and y axes. 

```{r resid-v-pred}
ggplot(data = cats, aes(x = Bwt, y = resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "_____", y = "______", 
       title = "________")
```

- What assumption(s) can you assess from this scatterplot? Are these assumptions satisfied? Briefly explain. 


5. Plot a histogram and Normal QQ plot of the residuals. Refer to [this slide](https://www2.stat.duke.edu/courses/Fall19/sta210.001/slides/lec-slides/04-slr-inf.html#27) for code to make a Normal QQ plot. 

```{r resid-histogram}

```

```{r resid-qqplot}


```

- What assumption(s) can you assess from the histogram and Normal QQ-Plot? Are these assumptions satisfied? Briefly explain. 

6. Comment on any other assumptions that haven't been addressed and whether they are satisfied. 


## Part 3: Statistical Inference

### Confidence Intervals 

7. Write the 95% confidence interval for the slope of body weight and interpret it in the context of the data. 


8. Based on the interval, do you think there is a statistically significant relationship between a cat's body weight and heart weight? Briefly explain. 

### Hypothesis Tests
 
9. State the hypotheses in (1) words and (2) statistical notation.  

10. What is the test statistic? The p-value? (Refer to the model output in a previous problem. )

11. State the conclusion in context of the problem.

12. Do the conclusions drawn from the hypothesis test and confidence interval agree? Briefly explain. 



## Part 4: Prediction 

13. Calculate the 95% confidence interval for the predicted heart weight of an individual cat that weighs 3.18 kg. Interpret the interval in context.

```{r}
x0 <- data.frame(Bwt = c(3.18))
predict.lm(bwt_hwt_model, x0, interval = "confidence", 
           conf.level = 0.95) 
```


14. Calculate a 95% prediction interval for the predicted mean heart weight for the subset of cats who weigh 3.18 kg. Interpret the interval in context.

```{r}
x0 <- data.frame(Bwt = c(3.18))
predict.lm(bwt_hwt_model, x0, interval = "prediction", 
           conf.level = 0.95) #<<
```