---
title: "Wages"
subtitle: "Starting salaries at a bank between 1965 and 1975"
date: "`r Sys.Date()`"
output: html_document
---

```{r global-options, echo = FALSE}
knitr::opts_chunk$set(warning=FALSE,
                   message=FALSE)
```

```{r load-packages}
library(tidyverse)
library(knitr)
library(broom)
library(Sleuth3) #case1202 data
library(cowplot) #plot_grid function
```

```{r prepare-data}
wages <- case1202 %>% 
  mutate(Female = as.factor(ifelse(Sex=="Female",1,0))) %>%
  select(-Sal77,-Sex)
```

```{r scatterplot}
pairs(Bsal ~ Senior + Age + Educ + Exper, data = wages, 
      lower.panel = NULL)
```

### Initial model

```{r first_model}
bsal_model <- lm(Bsal ~ Senior + Age + Educ + Exper + Female, 
            data=wages)
tidy(bsal_model,conf.int=TRUE)
```

### Checking Model Assumptions

```{r calc-resid}
wages <- wages %>% 
  mutate(predicted = predict.lm(bsal_model), residuals = resid(bsal_model))
```

```{r resid-vs-pred}
ggplot(data=wages,aes(x=predicted, y=residuals)) + 
  geom_point() + 
  geom_hline(yintercept=0,color="red") +
  labs(title="Residuals vs. Predicted Values") 
```

```{r resid-vs-x}
p1 <- ggplot(data=wages,aes(x=Senior,y=residuals)) + 
  geom_point() + 
  geom_hline(yintercept=0,color="red") +
  labs(x = "Seniority", y="Residuals")

p2 <- ggplot(data=wages,aes(x=Age,y=residuals)) + 
  geom_point() + 
  geom_hline(yintercept=0,color="red") +
  labs(x = "Age", y="Residuals")

p3 <- ggplot(data=wages,aes(x=Educ,y=residuals)) + 
  geom_point() + 
  geom_hline(yintercept=0,color="red") +
  labs(x = "Education", y="Residuals")

p4 <- ggplot(data=wages,aes(x=Exper,y=residuals)) + 
  geom_point() + 
  geom_hline(yintercept=0,color="red") +
  labs(x = "Experience", y="Residuals")

plot_grid(p1,p2,p3,p4,nrow=2)
```

```{r resid-vs-x-categorical}
ggplot(data=wages,aes(x=Female,y=residuals)) + 
  geom_boxplot() + 
  geom_hline(yintercept=0,color="red") +
  labs(x = "Female", 
       y="Residuals")
```

```{r resid-hist}
ggplot(data=wages,aes(x=residuals)) + geom_histogram(fill="steelblue2",color="black") + 
  labs(title="Distribution of Residuals") + 
    theme(plot.title=element_text(hjust=0.5,size=16))
```

```{r resid-qq}
ggplot(data=wages,aes(sample=residuals)) + stat_qq() + stat_qq_line() +
  labs(title="QQ-Plot of Residuals") + 
    theme(plot.title=element_text(hjust=0.5,size=16))
```


### Model with mean-centered variables

```{r}
wages <- wages %>%
  mutate(SeniorCent = Senior - mean(Senior), 
         AgeCent = Age-mean(Age), 
         EducCent = Educ - mean(Educ), 
         ExperCent = Exper - mean(Exper))
```

- Fit the regression model using the mean-centered variables. 

```{r}

```

- How did the model stay the same?
- How did the model change?

### Model with indicator variables

- Use the code below to create a categorical variable for `Educ`.

```{r}
wages <- wages %>%
  mutate(EducCat = as.factor(Educ))
```

- Fit the regression model using `EducCat` instead of `Educ`.

```{r}

```

- What is the baseline for Education?

- Interpret the coefficient for `EducCat16`.

- What is your conclusion from the p-value of `EducCat16`?

- Write the model equation for those with 8 years of education.

- Write the model equation for those with 16 years of education. 
