---
title: "Multinomial Logistic Regression"
subtitle: "*Sesame Street* Analysis"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE,
	echo= TRUE
)
```

```{r,echo=F}
library(tidyverse)
library(knitr)
library(broom)
library(cowplot)
library(arm) #binned residuals
library(nnet) #multinomial regression models
```

```{r}
# read in dataset
sesame <- read_csv("data/sesame.csv")
```

```{r}
# mean-center relevant continuous variables, make categorical variables factors
sesame <- sesame %>% 
  mutate(viewcat = as.factor(viewcat), 
         site = as.factor(site), 
         prenumbCent = prenumb - mean(prenumb), 
         preletCent = prelet - mean(prelet), 
         ageCent = age - mean(age),
         viewenc = ifelse(viewenc == 1, "Encouraged", "Not Encouraged"))
```
<br>

### Questions

1. Describe how you would conduct exploratory data analysis. What plots and/or summary statistics would you include? 

```{r warning = F, message = F}
model1 <- multinom(viewcat ~ site + viewenc + prenumbCent + preletCent + ageCent,
                   data = sesame)
kable(tidy(model1, conf.int=TRUE, exponentiate = FALSE),
      format = "markdown")
```
<br>



2. Interpret the intercept for the part of the model of the odds of `viewcat == 2` versus `viewcat == 1`. 

3. Interpret the effect of the numbers pretest on the odds of viewership. 

4. The primary objective of the experiment was to understand the effect of encouragement `viewenc` on viewership. Does encouragement have a significant effect on viewership? If so, describe the effect. Otherwise, explain why not. 

5. We want to test if there are any significant interactions with `viewenc` and the pretests. We create a model that includes the variables from `model1` along with `viewenc*preletCent` and `viewenc*prenumbCent`. 

```{r echo = T, warning = F, message = F}
model2 <- multinom(viewcat ~ site + viewenc + prenumbCent + preletCent + ageCent + 
                     viewenc*preletCent + viewenc*prenumbCent,
                   data = sesame)
```

The results from the drop-in-deviance test are shown below. Is there evidence of a significant interaction effect? Explain. 

```{r}
anova(model1, model2, test = "Chisq")
```

6. Describe the plots, tables, and/or calculations you would create using the residuals to assess model fit.