---
title: "Multinomial Logistic Regression"
subtitle: "*Sesame Street* Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height =5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE,
	echo= TRUE
)
```

```{r,echo=F}
library(tidyverse)
library(knitr)
library(broom)
library(cowplot)
library(nnet) #multinomial regression models
```

The main objective of this analysis is to understand how encouragement affects the frequency that children watch *Sesame Street*. We will use the following variables: 

**Response:**

- <font class="vocab">`viewcat`</font>
    + 1: rarely watched show
    + 2: once or twice a week
    + 3: three to five times a week
    + 4: watched show on average more than five times a week

**Predictors:**

- <font class="vocab">`age`:</font> child's age in months
- <font class="vocab">`prenumb`: </font>score on numbers pretest (0 to 54)
- <font class="vocab">`prelet`: </font>score on letters pretest (0 to 58)
- <font class="vocab">`viewenc`:</font> 1: encouraged to watch, 2: not encouraged
- <font class="vocab">`site:`</font>
  + 1: three to five year old from urban area
  + 2: four year old from suburban area
  + 3: from rural area with high socioeconomic status
  + 4: from rural area with low socioeconomic status
  + 5: from Spanish speaking home

```{r}
# read in dataset
sesame <- read_csv("data/sesame.csv")
```

```{r}
# mean-center relevant continuous variables, make categorical variables factors
sesame <- sesame %>% 
  mutate(viewcat = as.factor(viewcat), 
         site = as.factor(site), 
         prenumbCent = prenumb - mean(prenumb), 
         preletCent = prelet - mean(prelet), 
         ageCent = age - mean(age),
         viewenc = ifelse(viewenc == 1, "Encouraged", "Not Encouraged"))
```
<br>

### Questions


1. We'd like to fit a model to predict how often a child in this study watched *Sesame Street*. What type of model should we fit? Briefly explain your choice. 

#### Exploratory data analysis

2. Describe how you would conduct exploratory data analysis. What plots and/or summary statistics would you include? What information would you learn from the exploratory data analysis?

3. Make a plot to examine the relationship between the response and `viewenc`, the primary variable of interest in this analysis. What do you learn from the plot?

4. Make a plot to examine the relationship between the response and a quantitative predictor variable. What do you learn from the plot?

#### Model selection

5. The primary objective of the experiment was to understand the effect of encouragement, `viewenc`, on viewership. Explain the process you would use to select the final model. 

6. Conduct model selection and display the final model.

7. Interpret the intercept associated with the odds of a child being in the category `viewcat == 2` versus `viewcat == 1`. 

8. Interpret the coefficient one numeric predictor in terms of the odds of a child being in the category `viewcat == 2` versus `viewcat == 1`. Based on the confidence interval for the coefficient, is the numeric predictor a statistically significant predictor of viewership?

9. The primary objective of the experiment was to understand the effect of encouragement `viewenc` on viewership. Does encouragement have a significant effect on viewership? If so, describe the effect. Otherwise, explain why not.

10. We want to test if there is a statistically significant interaction between `viewenc` and `site`. Use the appropriate test to determine if this interaction is significant and should be included in the model. 

#### Model Assessment

11. How would you assess the appropriateness of the model fit? Describe the plots, tables, and/or calculations you would create to assess model fit. 

12. How would you assess the model's power in predicting the a child's viewership? Use the appropriate plot, table, and/or calculation to make a conclusion. 

### Solutions

https://www2.stat.duke.edu/courses/Fall19/sta210.001/appex/21-multinom-logistic-solutions.html

### References

- Summary of the original study: [A Summary of the Major Findings in "The First Year of
Sesame Street: An Evaluation"](https://files.eric.ed.gov/fulltext/ED122799.pdf)

- Data from [http://www2.stat.duke.edu/~jerry/sta210/sesamelab.html](http://www2.stat.duke.edu/~jerry/sta210/sesamelab.html)

