---
title: "Exam 02 Review"
date: "`r Sys.Date()`"
output: html_document
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                   message=FALSE)
```

```{r load-packages}
library(Sleuth3)
library(knitr)
library(broom)
library(tidyverse)
library(nnet)
library(cowplot)
```
```{r echo = F}
library(Sleuth3)
claws <- ex0722
claws <- claws %>% mutate(species_int = case_when(
  Species=="Cancer productus" ~ 1, 
  Species =="Hemigrapsus nudus" ~ 2, 
  Species =="Lophopanopeus bellus" ~ 3
), 
cp = factor(if_else(Species=="Cancer productus",1,0)),
       hn = factor(if_else(Species=="Hemigrapsus nudus",1,0)),
       lb = factor(if_else(Species=="Lophopanopeus bellus",1,0)), 
Species =  case_when(
  Species=="Cancer productus" ~ "Cp", 
  Species =="Hemigrapsus nudus" ~ "Hn", 
  Species =="Lophopanopeus bellus" ~ "Lb"
), 
forceCent = Force - mean(Force), 
heightCent = Height - mean(Height)
)
```

### Data Description

- We would like to identify crab species based on the closing force and propodus height of claws
  + `ex0722` data set in the `Sleuth3` R package

- **Predictors:** 
  + `Force`: </font> Closing force of claw (newtons)
  + `Height`: </font> Propodus height (mm)
  
- **Response:**
  + `Species`: </font> Hemigrapsus nudus (Hn), Lophopanopeus bellus (Lb), Cancer productus (Cp)

### Part I: Lb species?

Suppose we want to use `Force` and `Height` to determine whether or not a crab is from the Lophopanopeus bellus (Lb) species. 


**1. What type of model should we use? Briefly explain your choice.**

   <font color = "blue">Logistic regression, because the response is categorical with two levels.</font>

**2. Briefly explain how you would conduct exploratory data analysis.** 
 
  <font color = "blue">Below are some plots that can be used to conduct exploratory data analysis.</font>
  
```{r}
p1 <- ggplot(data = claws, aes(x = Height)) +
  geom_histogram() + 
  labs(title = "Distribution of Height", 
       x = "Height (in mm)")

p2 <- ggplot(data = claws, aes(x = Force)) +
  geom_histogram() + 
  labs(title = "Distribution of Force", 
       x = "Force (in Newtons)")

p3 <- ggplot(data = claws, aes(x = lb)) +
  geom_bar() + 
  labs(title = "Distribution of Lb vs. Not Lb")

top_row <- plot_grid(p1, p2, ncol = 2)

plot_grid(top_row, p3, ncol = 1)
```

```{r}
p1 <- ggplot(data = claws, aes(x = lb, y = Force)) +
  geom_boxplot() + 
  labs(title = "Lb vs. Force")

p2 <- ggplot(data = claws, aes(x = lb, y = Height)) +
  geom_boxplot() + 
  labs(title = "Lb vs. Height")

plot_grid(p1, p2, ncol = 2)
```


We will use the mean-centered variables for `Force` and `Height`. The model output is below.

```{r echo = F}
m2 <- glm(lb ~ forceCent + heightCent, 
          data=claws, family = "binomial")
kable(tidy(m2),format="markdown",digits=3)
```


**3. Write the equation for the odds of a crab being from the Lb species.**

$$\color{blue}{\log\Big(\frac{\hat{p}}{1 - \hat{p}}\Big) = - 1.130 + 0.211 \times \text{ForceCent} - 0.895 \times \text{heightCent}}$$

**4. Interpret the intercept in the context of the problem.**

  <font color = "blue">The odds that a crab with average claw force (`r round(mean(claws$Force),3)`) and average claw height (`r round(mean(claws$Height),3)`) is the Lb species is exp(-1.130) = `r round(exp(-1.130), 3)`.</font>

**5. Interpret `forceCent` in the context of the problem.**

  <font color = "blue">For each additional Newton in force, we expect the odds that a crab is the Lb species to multiply by exp(0.211) = `r round(exp(0.211), 3)` holding height constant.</font>

The ROC curve is below. 

```{r echo = F, fig.height = 3, fig.width = 4}
claws_aug <- augment(m2, type.predict = "response", 
                     type.residuals = "response")

library(plotROC) #extension of ggplot2
roc_curve <- ggplot(claws_aug, aes(d = as.numeric(lb), m = .fitted)) +
  geom_roc(n.cuts = 5, labelround = 3) + 
  geom_abline(intercept = 0) + 
  labs(x = "False Positive", 
      y = "True Positive")
roc_curve
calc_auc(roc_curve)$AUC
```


**6. What does <i>sensitivity</i> mean in the context of this data? What does <i>specificity</i> mean?**

  <font color = "blue">Sensitivity: Correctly predicting a crab is the Lb species.</font>
  <font color = "blue">Specificity: Correctly predicting a crab is not in the Lb species.</font>
    
**7. Suppose we use a threshold of 0.507. What is the <i>sensitivity</i> at this threshold? What is the specificity?** 

  <font color = "blue">Sensitivity is about 0.5, and specificity is about 0.85.</font>

### Part 2: Which species?

Suppose we want to use force and height to determine a crab's species. The model output is below:


```{r, include=F}
m1 <- multinom(Species ~ forceCent + heightCent, data=claws)
```

```{r echo = F}
tidy(m1, exponentiate = FALSE) %>%
  kable(format = "markdown", digits = 3)
```

**1. Write the equation of the model.** 

$$\color{blue}{\log\Big(\frac{\hat{p}_{Hn}}{\hat{p}_{Cp}}\Big) = -1.193 - 0.494 \times \text{ForceCent} + 0.179 \times \text{heightCent}}$$
$$\color{blue}{\log\Big(\frac{\hat{p}_{Lb}}{\hat{p}_{Cp}}\Big) = 0.021 + 0.095 \times \text{ForceCent} - 0.902 \times \text{heightCent}}$$

**2. Interpret the intercept for the odds a crab is Hn vs. Cp species.**

  <font color = "blue">The odds that a crab with average claw force (`r round(mean(claws$Force),3)`) and average claw height (`r round(mean(claws$Height),3)`) is the Hn species vs. Cp species is exp(-1.193) = `r round(exp(-1.193), 3)`.</font>

**3. Interpret the coefficient of `forceCent` for the odds a crab is Lb vs. Cp species.**

  <font color = "blue">For each additional Newton in force, we expect the odds a crab is the Lb species vs. Cp species to multiply by exp(0.095) = `r round(exp(0.095), 3)` holding height constant.</font>