---
title: "Lab 01"
subtitle: "R Review"
date: "01.17.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

## Meet Your TA!

---

## Introduce Yourself!

- Name

- Major or intended major

- Share a boring fact about yourself 

---

## Today's Agenda

1. Questions about the syllabus 

2. What to expect in lab

3. Lab 01 intro and demos: Introduce the lab, and work through the first section as a class.

4. On your own: Work on the rest of the lab "on your own", but feel free to check in with classmates as much as you like.

---

### Any questions about the syllabus?

Course website: [http://bit.ly/sta210-sp20](http://bit.ly/sta210-sp20)

---

## What to expect in lab

- Introduction to the lab assignment (~ 5 - 10 min, longer today)

- Work on the lab assignment (individual today, but generally done in teams)

- Lab slides and instructions posted in the schedule on the course website under <i class="fa fa-file-code-o fa-lg"></i>
    + Go to the course website to find today's lab instructions.

- All assignments will be completed using RStudio and GitHub
    + You will submit your GitHub repo on Gradescopes

---

## Tips

- You do not have to finish the lab in class, they will always be due the following Tuesday at 11:59p. One work strategy is to get through portions that you think will be most challenging (which initially might be the coding component) during lab when a TA can help you on the spot and leave the narrative writing until later.

- Do not pressure each other to finish early; use the time wisely to really learn the material and produce a quality lab report.

---

class: center, middle

## Lab 01 intro and demos

---

## Goals

<small>
- **Introduce you to R and RStudio:**
    + R is the name of the programming language itself 
    + RStudio is a convenient interface
    + Access RStudio through the Docker container on Duke VM Manage - UPDATE LINK!


- **Introduce you to Git and GitHub:** collaboration and version control system that we will be using throughout the course
    + Git is a version control system -- like "Track Changes" features from Microsoft Word on steroids
    + GitHub is the home for your Git-based projects on the internet
    
- **Take small steps:**
    + This a solo lab to make versioning (working with Git) simpler
    + Next week you’ll learn about collaborating on GitHub and produce a single lab report for the team

</small>

---

## Getting started - CHANGE THESE INSTRUCTIONS!

You may watch or follow along: 

- Find the GitHub repo with the prefix `lab-01-review-r`. 

- On GitHub, click on the green **Clone or download button**, select **Use HTTPS** if it is not already selected. Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the sta210-fa19 course workspace. Create a **New Project from Git Repo**. You will need to click on the down arrow next to the **New Project** button to see this option.

- Copy and paste the URL of your assignment repo into the pop up dialog box.

- Click OK.

Now it's your turn! When you're done with this part, you may continue if you like or wait until I demonstrate the next part. Please raise your hand if you have any questions.

---

## Configure git - DO I STILL NEED THIS STEP??

Next, we need to configure git so RStudio and GitHub can communicate with each other. 

You may watch or follow along: 

- Type the following lines of code in the **console**: 

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="your name", user.email="your email")
```

- If you get an error message, it may be because you need to install the `usethis` package. You can type the code below in the console to install the package

```{r eval=FALSE}
install.packages("usethis")
```

Once you're configured git, you're ready to complete the rest of lab!\

---

### Submitting Assignments 

- Fill in these instructions! 

---

class: middle, center

## You're now ready to complete the rest of lab!

.instructions[
Please raise your hand if you need help as you work on the lab.
]

---

### Need more practice with R? - UPDATE THIS! 

- **Duke Libraries Rfun - Intro to R Workshop**: Data Transformations, Data Structures, and the Tidyverse
  - September 12 1p - 3p
  - To register: https://duke.libcal.com/event/5497129

- **RStudio Cloud Primer**: Work with Data

https://rstudio.cloud/learn/primers/2

- **R for Data Science**: Data Visualization

https://r4ds.had.co.nz/data-visualisation.html
