---
title: "Lab 01"
subtitle: "Intro to R"
date: "01.17.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Welcome to lab

---

## Agenda

1. **Lab strucuture:** Lab structure overview.

2. **Lab 01 intro and demos:** Introduce the lab, and work through the first section as a class.

3. **On your own:** Work on the rest of the lab "on your own", but feel free to check in with teammates as much as you like.

---

class: center, middle

# Lab strucuture

---

## Lab strucuture

- ~ 10 minute introduction (a bit longer today)

- Use the remaining time to work through the lab exercises and fill out your lab report
    + Today: on your own
    + Remainder of the semester: in teams

- Lab instructions posted on the course website under <i class="fa fa-file-code-o fa-lg"></i>
    + Go to the course website to find today's lab instructions.
    
---

## Tips

- You do not have to finish the lab in class, they will always be due the following Wednesday at 11:59p. One work strategy is to get through portions that you think will be most challenging (which initially might be the coding component) during lab when I can help you on the spot and leave the narrative writing till later.

- Do not pressure each other to finish early; use the time wisely to really learn the material and produce a quality report.

- When working in teams, do **not** split up the lab among the team members; work on it together in its entirety.

---

class: center, middle

# Lab 01 intro and demos

---

## Goals

<small>
- **Introduce you to R and RStudio:**
    + R is the name of the programming language itself 
    + RStudio is a convenient interface


- **Introduce you to Git and GitHub:** collaboration and version control system that we will be using throughout the course
    + Git is a version control system -- like "Track Changes" features from Microsoft Word on steroids
    + GitHub is the home for your Git-based projects on the internet
    
- **Take small steps:**
    + This a solo lab to make versioning (working with Git) simpler
    + Next week you’ll learn about collaborating on GitHub and produce a single lab report for the team

</small>

---

## Getting started

<small>
You may watch or follow along: 

- Find the GitHub repo with the prefix `lab-01-intro-r`. 

- On GitHub, click on the green **Clone or download button**, select **Use HTTPS** if it is not already selected. Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the course workspace. Create a **New Project from Git Repo**. You will need to click on the down arrow next to the **New Project** button to see this option.

- Copy and paste the URL of your assignment repo into the pop up dialog box.

- Click OK.

Now it's your turn! When you're done with this part, you may continue if you like or wait until I demonstrate the next part. Please raise your hand if you have any questions.
</small>

---

## Configure git

<small>
Next, we need to configure git so RStudio and GitHub can communicate with each other. 

You may watch or follow along: 

- Type the following lines of code in the **console**: 

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="your name", user.email="your email")
```

- If you get an error message, it may be because you need to install the `usethis` package. You can type the code below in the console to install the package

```{r eval=FALSE}
install.packages("usethis")
```

Once you're configured git, you're ready to complete the rest of lab!
</small>

---

class: middle, center

## You're now ready to complete the rest of lab!

---

## More practice with R (optional)

- RStudio Cloud Primer: Work with Data
    - [https://rstudio.cloud/learn/primers/2](https://rstudio.cloud/learn/primers/2)

- DVS Intro to R Workshop - Jan 22 at 2p
    - [https://duke.libcal.com/event/4799048](https://duke.libcal.com/event/4799048)
 
- R for Data Science: Data Visualization
    - [https://r4ds.had.co.nz/data-visualisation.html](https://r4ds.had.co.nz/data-visualisation.html)



