---
title: "Lab 02"
subtitle: "Simple Linear Regression"
author: ""
date: "01.23.20"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Welcome to lab

---

## Agenda

1. **Teams:** Find teammates.

2. **Tips:** Working in RStudio & GitHub

3. **Starting a new project:** Clone the repo into new Rstudio project

4. **Configure git**: Make sure your git config is set up

5. **Warm-up:** Make sure you know how to push and pull.


---
    
## Teams

- Find your teammates: [STA210 Teams](https://prodduke-my.sharepoint.com/:x:/g/personal/mt324_duke_edu/EZaLCjdJR7hKhHScho6x66IB-p48xcaVwzMy2-0G2e3xKg?e=HQglFh)

- By the end of today's lab: 
    - Come up with a **team name**
    - Find a **weekly 2-hour block between Thursday and Tuesday (outside of the lab session)** that the team can meet to finish lab. You may not need to use the time every week, but it will be helpful to already have a plan for the weeks that you do need extra time.
    - Come up with away to communicate outside of labs and group meetings (email, text, Slack, etc.).
    - [Click here](https://forms.gle/kTz8NKEDeVj28hSX9) to report your team name, weekly meeting time, and primary form of group communication (one per)
    
---

## Working on a team

- Do not pressure each other to finish early; use the time wisely to really learn the material and produce a quality report.

- The labs are structured to help you learn the steps of a regression analysis. Do not split up the lab among the team members; work on it together in its entirety.

- Everyone has something to contribute! Use the lab groups as an opportunity to share ideas and learn from each other.

---

### Starting a new project

You may watch or follow along: 

- Find the GitHub repo with the prefix `lab-02-`. 

- On GitHub, click on the green **Clone or download button**, select **Use HTTPS** if it is not already selected. Click on the clipboard icon to copy the repo URL.

- Go to https://vm-manage.oit.duke.edu/containers and login with your Duke NetId and Password.

- Click to log into the Docker container **STA 210 - Regression Analysis**. You should now see the RStudio environment. 

---

### Starting a new project

- Go to *File* `r emo::ji("arrow_right")` *New Project* `r emo::ji("arrow_right")` *Version Control* `r emo::ji("arrow_right")` *Git*. 

- Copy and paste the URL of your assignment repo into the dialog box *Repository URL*. You can leave the direct name empty. It will default to the name of the GitHub repo.


- Click *Create Project*, and you should see the contents from your GitHub repo in the **Files** pane in RStudio. 

---

## Configure Git

- If you need to configure git, type the following lines of code in the console in RStudio filling in your name and email address.

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="your name", user.email="your email")
```

- Before starting on lab, let's talk about the workflow now that you're on a team

---

## Team Workflow

- You and your teammates are working from the same lab-02 GitHub repo. This means you are all contributing to the same R Markdown document!

- For today's lab, **<i>only one person should be typing the code/narrative in the R Markdown document at at time</i>**. 
    - Rotate who updates the document based on the markers in the lab instructions. **<i>Each team member should have at least one commit in the lab.</i>**
    - When you're not typing, you should still be contributing to the discussion but do not actually touch the files on your computer.

---

### Team Workflow

I will now demonstrate the steps for updating the lab documents and sharing the updated files with your teammates. 

- **Every team member**: Click the **Pull** button in the Git pane. This brings the most updated files from GitHub to your RStudio project. 

- **<u>ONE</u> team member**:
  - Update the .Rmd file.
  - Knit the .Rmd file. 
  - Click the box next to <u>every</u> file in the Git pane.
  - Write a short and informative commit message. 
  - Push the changes to GitHub.

- **Every team member**: Click the **Pull** button in the Git pane. You should now see the changes reflected in the documents in your RStudio project.

---

## Warm-up

- Complete the warmup exercise to practice pushing and pulling changes.

- Once you've completed the warmup read through the tips on the next slide. You are then ready to start working on the rest of lab.
    - **Remember: only one person should updating the .Rmd file during the warmup**
    - We will talk more about working with teams in GitHub next week.

- Please raise your hand if you have any questions about the warmup exercise. 


---

## Lab Tips

- Be ready to troubleshoot your document, since it will likely fail to knit on multiple occasions throughout the process. Read the error message carefully and take note of which line is preventing a successful knit.

- Make sure to keep track of your various chunks and to keep text and code in the right place.

- Remember that your R Markdown file is not aware of your project's global environment and can only make use of variables, functions, etc. that you have loaded or defined in the document.

- If you're unsure how a function works or what its arguments are, simply type `?` in front of it and hit enter (`?read_csv` for instance). The "Help" tab will open and provide a summary of the function as well as some examples.

---

---

### Submitting Assignments

- You will be submitting your final assignment in [Gradescope](http://www.gradescope.com)

- We will be grading the **PDF** in your GitHub repo, so make sure it has all of your final code, output, graphs, and narrative. 

- Detailed instructions on how to submit the lab are at the [end of the lab instructions](https://www2.stat.duke.edu/courses/Spring20/sta210.001/labs/lab-02-slr.html#submitting-the-assignment).


---

### Need more practice with R? 

- [Chapter 2: Get Started](https://socviz.co/gettingstarted.html#work-in-plain-text-using-rmarkdown) *Data Visualization by Kieran Healy* 

- [Chapter 3: Data visualization](https://r4ds.had.co.nz/data-visualisation.html) in *R for Data Science* by Hadley Wickham

- RStudio Cloud Primers
  - Visualization Basics: https://rstudio.cloud/learn/primers/1.1
  - Work with Data: https://rstudio.cloud/learn/primers/2
  - Visualize Data: https://rstudio.cloud/learn/primers/3

- Visualization in R using ggplot2 workshop by the Center for Data and Visualization Sciences 
  - Jan 30, 1p - 3p
  - More info & registration: https://duke.libcal.com/event/6085379

