---
title: "Lab 03: ANOVA"
author: ""
date: "02.01.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Welcome to lab

---

## Agenda

1. Team Name

2. Tips for collaborating 

3. Lab Introduction

4. Starting a new project & configure git

---

## Team Name

.middle[
If you haven't already, please give me your team name by the end of today's lab.
]

---

## Tips for collaborating via GitHub

- Always pull first before you start working.

- Knit, commit, and push often to minimize merge conflicts and/or to make merge conflicts easier to resolve.

- If you find yourself in a situation that is difficult to resolve, ask questions asap, don't let it linger and get bigger.

---

## Lab introduction 

- <font class="vocab">Goal: </font> Use ANOVA to compare group means 

- <font class="vocab">Data:</font> `diamonds` data set in the ggplot2 package  


- Learn about attributes of diamonds: 
    - [https://www.gia.edu/diamond-quality-factor](https://www.gia.edu/diamond-quality-factor)

- Today's lab will focus on color and carat weight
---

## Starting a new project

- Go to the course GitHub organization and find the **lab-03** repo that has your team name on it.

- On GitHub, click on the green Clone or download button, select Use HTTPS (this might already be selected by default, and if it is, you’ll see the text Clone with HTTPS). Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the course workspace. Create a New Project from Git Repo. You will need to click on the down arrow next to the New Project button to see this option.

- Copy and paste the URL of your assignment repo into the dialog box and click OK. 

---

## Configure Git

- Type the following lines of code in the console in RStudio filling in your name and email address.

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="your name", user.email="your email")
```

- Update the name of your RStudio project to "Lab 03 - ANOVA"

