---
title: "Lab 07:"
subtitle: "Model Selection & Diagnostics"
author: ""
date: "03.05.20"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


### Agenda


1. Mid-semester survey

2. Lab Introduction

---

### Mid-semester Survey

Please take a few minutes to fill out a short mid-semester survey: 

https://duke.qualtrics.com/jfe/form/SV_bQlQ0Z1VhOhZOD3

Your responses are anonymous.

---

## Lab introduction 

- <font class="vocab">Goal: </font> Better understand model selection and diagnostics and how to do these procedures in R
    - **Part I**: This lab will focus on model selection in the `regsubsets` and `step` functions using Adj. $R^2$, BIC, and AIC as selection criteria
    - **Part II**: Model diagnostics (leverage, standardized residuals and VIF) 

- <font class="vocab">Tips: </font>
    - Use the [Model Selection & Diagnostics notes](https://www2.stat.duke.edu/courses/Spring20/sta210.001/slides/lec-slides/12-model-selection.html) to help you complete the lab. Some of the primary functions you'll use are on the next two slides.
    
---

### `step` function (AIC)

```{r,eval=F}
null_model <- lm(Y ~ 1, data = my_data)
full_model <- lm(Y ~ ., data = my_data)
```

- Forward selection 

```{r eval=F}
regfit_forward <- step(null_model, scope = formula(full_model), 
                       direction = "forward")
```


- Backward selection
```{r,eval=F}
regfit_backward <- step(full_model, direction = "backward")
```


---

### `regsubsets` function (BIC, Adj. $R^2$)

```{r eval = F}
library(leaps)
```

- Forward selection

```{r eval=F}
regfit_forward <- regsubsets(Y ~ ., data = my_data, method="forward")
```

- Backward selection 

```{r eval=F}
regfit_backward <- regsubsets(Y ~ ., data = my_data, method="backward")
```

- Select a model

```{r eval=F}
sel_summary <- summary(regfit_forward)
coef(regfit_backward, which.max(sel_summary$adjr2)) # Adj R-sq 
coef(regfit_backward, which.min(sel_summary$bic)) # BIC 
```

---

### Project

- If you finish the lab early, you should use the remaining time to work on the project proposal 

- Details about the proposal may be found in the [project instructions](https://www2.stat.duke.edu/courses/Spring20/sta210.001/project/project.html)

- Please let your TA know if you have any questions about the proposal or finding data


