---
title: "Lab 07:"
subtitle: "Logistic Regression, Part 1"
author: ""
date: "10.23.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


### Agenda


1. Lab introduction

2. Getting started

3. Project

---

### Lab 07: Logistic Regression

<font class="vocab">Goal: </font> Fit a logistic regression model to predict the probability a Spotify user will like a song
- This week will focus on exploratory data analysis and model fit. Next week, you will assess the fit of the model and use it to predict if a user likes selected songs

<font class="vocab">Data: </font> Song characteristics from Spotify.com and like/dislike from a Spotify user
- You will download the data from the lab instructions and upload it into the data folder of your project. 
- Links to the dataset are in the Data section of the instructions.

**Use the lecture notes on logistic regression for help with coding and interpretations.**


---

### Getting started with the lab

- Go to the course GitHub organization and find the **lab-07** repo that has your team name on it.

- On GitHub, click on the green Clone or download button, select Use HTTPS (this might already be selected by default, and if it is, you’ll see the text Clone with HTTPS). Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the course workspace. Create a New Project from Git Repo. You will need to click on the down arrow next to the New Project button to see this option.

- Copy and paste the URL of your assignment repo into the dialog box and click OK. 

- Run the following in the console to configure Git:

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="github username", 
               user.email="your email")
```

- **Use the email address associated with your GitHub account**

---

### Caching your password

You need to cache your password on a per-project basis: 

- Go to the **Terminal** within a project (the most recent lab?)
- Type the following to cache your password for $60 * 60 * 24 * 7 = 604800$ seconds

```{bash eval=FALSE}
git config --global credential.helper 'cache --timeout 604800'
```

You will need to enter your GitHub username and password one more time after caching the password. After that you won't need to enter your credentials for 604800 seconds = 7 days.

---

### Project

- If you finish the lab early, you should use the remaining time to work on the project proposal 

- Details about the proposal may be found in the [project instructions](https://www2.stat.duke.edu/courses/Fall19/sta210.001/project/project.html)

- Please let your TA know if you have any questions about the proposal or finding data


