---
title: "Lab 08"
subtitle: "Logistic Regression, Part 2"
author: ""
date: "Halloween 2019 `r emo::ji('jack_o_lantern')`"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


### Agenda


1. Lab introduction

2. Getting started

---

### Lab 07: Logistic Regression Part 2

<font class="vocab">Goal: </font> Fit a logistic regression model to predict the probability a Spotify user will like a song
- This week's focus is on assessing the model fit and using it to predict whether a user likes a few selected songs. 

<font class="vocab">Data: </font> Song characteristics from Spotify.com and like/dislike from a Spotify user
- You will download the datasets `spotify.csv` and `test_songs.csv` from the lab instructions and upload them into the data folder of your project. 
- Link to the data dictionary are on the [Spotify documentation page](https://developer.spotify.com/documentation/web-api/reference/tracks/get-audio-features/)

**Use the lecture notes on logistic regression for help with coding and interpretations.**


---

### Getting started with the lab

- Go to the course GitHub organization and find the **lab-08** repo that has your team name on it.

- On GitHub, click on the green Clone or download button, select Use HTTPS (this might already be selected by default, and if it is, you’ll see the text Clone with HTTPS). Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the course workspace. Create a New Project from Git Repo. You will need to click on the down arrow next to the New Project button to see this option.

- Copy and paste the URL of your assignment repo into the dialog box and click OK. 

- Run the following in the console to configure Git:

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="github username", 
               user.email="your email")
```

- **Use the email address associated with your GitHub account**

---

### Caching your password

You need to cache your password on a per-project basis: 

- Go to the **Terminal** within a project (the most recent lab?)
- Type the following to cache your password for $60 * 60 * 24 * 7 = 604800$ seconds

```{bash eval=FALSE}
git config --global credential.helper 'cache --timeout 604800'
```

You will need to enter your GitHub username and password one more time after caching the password. After that you won't need to enter your credentials for 604800 seconds = 7 days.


