---
title: "Lab 08"
subtitle: "Logistic Regression"
author: ""
date: "04.02.20"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


### Lab grading

- **Completion**
  - **Complete**: Reasonable attempt at every question, work submitted as PDF, document, neatly organized with appropriate headings (50 pts)
   - **Partially Complete**: Reasonable attempt at most questions, work submitted as PDF, document, neatly organized with appropriate headings (25 pts)
  - **Incomplete**: Most questions unanswered or partially answered (0 pts)

- **Formatting**
  - Work not submitted as PDF (-5 pts)
  - Document not neatly organized, unclear or missing exercise headers (-5 pts)

---

### Tip for remote collaboration: GitHub issues

- Each GitHub repo has a tab called "Issues" that can be used to keep track of what needs to be done on a given project. 
  - GitHub issues are especially useful when working with a team remotely, 

- One way to use issues on lab assignments and the project: 
  - Create a new issue in the repo that includes a list of tasks for the assignment/project,
  - Assign each task to a team member by tagging their GitHub username (e.g. @Username)
  
- Learn more about creating a new issue [here](https://help.github.com/en/github/managing-your-work-on-github/creating-an-issue).

---

### Lab 08 Introduction

<font class="vocab">Goal: </font> Fit a logistic regression model to predict the probability a Spotify user will like a song

<font class="vocab">Data: </font> Song characteristics from Spotify.com and like/dislike from a Spotify user
- Link to the data dictionary are on the [Spotify documentation page](https://developer.spotify.com/documentation/web-api/reference/tracks/get-audio-features/)

- Need a reminder on the code? See [logistic.html](https://www2.stat.duke.edu/courses/Spring20/sta210.001/appex/logistic.html)

