---
title: "Lab 09: Putting it All Together"
author: ""
date: "04.05.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Agenda

1. Lab introduction

2. Goals for this lab

3. Getting started

---

### Lab 09: Putting it All Together

- <font class="vocab">Goals</font>:
    - Conduct a full statistical analysis and write results as a short report
    - Practice imputing missing values
    - Use k-fold cross validation to assess model performance
    
- <font class="vocab">Data</font>: `gapminder` dataset in the **dslabs** package

- Use the lecture notes on [missing data](https://www2.stat.duke.edu/courses/Spring19/sta210.001/slides/lec-slides/20-missing-data.html#1) and [k-fold cross validation](https://www2.stat.duke.edu/courses/Spring19/sta210.001/slides/lec-slides/19-validation.html#1) for help with concepts and code.

---

## Goals for this lab

- Part of the final project will be writing a report (up to 3 pages) about your analysis

- Though there are 3 - 4 people in each group, the final report should read as if it was written by a single author

- This lab will give your group practice putting together a short report before working on the final project. Use this lab to...
    - figure out a work flow that you can use when you work on the final project
    - figure out each group member's strengths and use this to help divide up work for the final project 
    
---

## Getting started with the lab

- Go to the course GitHub organization and find the **lab-09** repo that has your team name on it.

- On GitHub, click on the green Clone or download button, select Use HTTPS (this might already be selected by default, and if it is, you’ll see the text Clone with HTTPS). Click on the clipboard icon to copy the repo URL.

- Go to RStudio Cloud and into the course workspace. Create a New Project from Git Repo. You will need to click on the down arrow next to the New Project button to see this option.

- Copy and paste the URL of your assignment repo into the dialog box and click OK. 

- Run the following in the console to configure Git:

```{r eval=FALSE}
library(usethis)
use_git_config(user.name="your name", user.email="your email")
```

- **Use the email address associated with your GitHub account**

---

## Caching your password

You need to cache your password on a per-project basis: 

- Go to the **Terminal** within a project (the most recent lab?)
- Type the following to cache your password for $60 * 60 * 24 * 7 = 604800$ seconds

```{bash eval=FALSE}
git config --global credential.helper 'cache --timeout 604800'
```

You will need to enter your GitHub username and password one more time after caching the password. After that you won't need to enter your credentials for 604800 seconds = 7 days.



