---
title: "Project Work Day"
author: ""
date: "11.14.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

### Collaborative Learning in STEM

Please fill out the survey below. The data is for research conducted by the Collaborative Learning in STEM Bass Connections project.

Participation is voluntary.

https://duke.qualtrics.com/jfe/form/SV_8i6V0yoMDcpVWg5

---


### Project Work Day

- Today's lab is all about working on the regression analysis for your project.

- We will walk around throughout the lab section and talk with each group about their project. 

- Fill out this short form at the end of lab: https://forms.gle/BNpE1B4JPzqtp3KT8 .

