---
title: "STA 210 Project"
subtitle: "Find Data!"
author: ""
date: "02.27.20"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta210-slides.css"
    logo: img/sta210-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

---

### Project

- The primary task for today's lab is to start identifying a dataset your group would like to analyze for the STA 210 project.

-  You'll go over the project in more detail on Monday, but you can see an outline of the project instructions [here](https://www2.stat.duke.edu/courses/Spring20/sta210.001/project/project.html)

- You now have a GitHub repo with the prefix `project-`. This is where you will put all of the work for the final project. 


---

### Project ideas assignment 

You will briefly write about a dataset you'd like to use for the project in the file `proposal.Rmd`. This includes 

- **Narrative** 
  - What is the research question you wish to explore? What's your motivation for examining this question?
  - Describe the dataset you wish to explore. 

- **The Data**
  - Data dictionary in the README of the `data` folder
  - Import the data into the `data` folder and include a `glimpse` of the data

See the [project instructions](https://www2.stat.duke.edu/courses/Spring20/sta210.001/project/project.html) for more details 

---

### General data guidelines 

- It is best to start with the question of interest and finding the data second. In general, we are using a set of predictor variables to understand variations in a response variable.

- Your response variable can be **quantitative** <u>or</u> **categorical** 
    - We will talk about models for categorical response variables after spring break (in plently of time to do the project)!

- Your dataset must have at least 100 observations and at least 10 variables (exceptions can be made but you must speak with me first). 
  - The data set should include both quantitative and categorical variables.

---

### General data guidelines 

- If you want to analyze data that will almost certainly have violations of assumptions, e.g. data collected over time, you will need to think of ways to address these violations in the project.

-  As you’re looking for data, keep in mind your regression analysis must be done using R.
  - Once you find a data set, you should make sure you are able to load it into RStudio, especially if it is in a format we haven’t used in class before.

- **You may not reuse datasets used in in-class examples/homework/labs.**

---

### Project Option: Modeling attendance at a music venue in NC

- This semester, we have data from a music venue in North Carolina that includes information about attendance at events along with info about social media prior to the event and some band characteristics 

- **Goal**: Fit a regression model to help the venue understand which variables are most useful in understanding variation in attendance

---

### Project Option: Modeling attendance at a music venue in NC

- **Main Data Challenges**: 
  - *Serial Correlation*: the data is collected over time, so you will need to think about how to deal with this
  - *Correlated Predictors*: a lot of the predictors are related to social media, so you will have to consider how to maximize the information you can get from the data (we'll talk more about correlated predictors next week!)

- This data is good for groups who: 
  - Are interested in music and/or social media
  - Want to work on a messy but fun dataset! 

---

### Project Option: Modeling attendance at a music venue in NC

- At most 8 groups will be able to work on this project. 

- If you are interested in this project, please let me know by filling out [this form](https://docs.google.com/forms/d/e/1FAIpQLSe1P-DbbT6RSx9qVVktXwwFp6UsbesjuSZGmSSBYf6Th7s0MA/viewform?usp=sf_link)

- You will still need to complete the Proposal for another dataset even if you are interested in this project

---

### Submitting work for the project

- To give you some experience with the type of workflow you’ll experience in practice, you will submit the proposal and all other components of the project by pushing your work to the project GitHub repo.

- You do not need to submit anything for the project on Gradescope.

- We will provide comment and feedback as an “issue” in your GitHub repo (more on this later)





